---
title: "Functional enrichment analysis"
output: html_document
author: "J. Ignacio Lucas Lledó"
params:
   TAG: 'genes'
   VAR: 'regime'
bibliography: enrichment.bib
---

```{r setup, message=FALSE}
library(topGO)
library(knitr)   # to use kable()
library(limma)   # to use vennDiagram()
library(ggplot2)
library(variancePartition)
TAG <- params$TAG
VAR <- params$VAR
ANNOTATION <- list(genes = '../2019-07-26/genes/annotation.txt',
                   isoforms = '../2019-07-26/transcripts/annotation.txt')
DGEDIR <- '../2020-01-08/'
ENRICHMENT <- paste('../2020-01-14/Enrichment', TAG, VAR, 'RData', sep='.')
```

# Introduction
This is the enrichment analysis for `r TAG` regulated by `r VAR`. I am using the 
`topGO` package [@Alexa2019], which is able to apply different algorithms. The starting point
is an ordering of `r TAG` usually according to their association with a phenotype or
any other relevant quantity, such as a measure of differential expression between
two conditions. In this case, we order `r TAG` by their differential expression
between `r VAR` levels. Following one of the suggestions in the manual, I originally
ordered `r TAG` by either $p$ value of the differential expression analysis, or by
the (complement of the) amount of its expression variance explained by `r VAR`
(see `2020-01-14`). Those were quite equivalent orderings. However, they do not
distinguish the sign of the fold change: both up- and down-regulated genes contribute
to make a gene ontology term significant. It is one of the approaches recommended,
and it makes sense, to the extent that among all the genes annotated with a function,
some may display contrasting expression patterns, for example if both positive and
negative regulators of that function are annotated with the same label. I thus,
expect the use of $p$ values in enrichment analysis to facilitate the detection of
high level categories. 

In contrast, genes sharing lower level categories (very specific GO terms) 
are more likely to be expressed in a similar way. Then, an ordering of genes that
reflect the sign of the difference in expression levels between the two conditions
compared would be more informative. In principle, enrichment methods should be able
to detect an enrichment in either end of the list, rather than only in the top.
That is actually the case for tests based on the Kolmogorov-Smirnov distribution,
like GSEA [@Subramanian2005]. I actually checked that reversing the order of genes
produces the exact same results, as expected. Below, I use the Kolmogorov-Smirnov
test.

For gene ordering, I use the $t$ statistic of the differential expression analysis,
which is positive when expression is higher in the random than in the regular regime,
and vice versa.

# Reading the data
Functional annotation is in `2019-07-26`. I will also upload two lists of `r TAG`,
with either proportion of variance explained by `r VAR` or p-value of differential
expression test.

```{r data}
load(paste0(DGEDIR, TAG, '.RData'))
tagTStat    <- topTable(fitmm, coef = VAR, num = length(fitmm$F), sort.by='t', resort.by='t')[,'t',drop=FALSE]
annotation  <- read.table(ANNOTATION[[TAG]], col.names = c('tagname', 'goterms'))
```

To initialize the `topGOdata` object, I need the gene list as a named numeric vector,
where the names are the `r TAG` identifiers and the numeric values, the $t$ statistics
from the differential expression analysis. The `structure()` function adds attributes
to an object.

```{r scores}
TStats <- structure(tagTStat[,1], names = row.names(tagTStat))
rm(tagTStat)
```

There are `r length(TStats)` `r TAG` scored with a $t$ statistic.
The `annotation` data frame has more than one GO term for every tag, separated by `|`.
I need a named list.

```{r annotation}
head(annotation)
allgenes2GO <- strsplit(as.character(annotation$goterms), "|", fixed = TRUE)
names(allgenes2GO) <- annotation$tagname
rm(annotation)
```

There are `r length(allgenes2GO)` `r TAG` with GO annotations. But the differential expression
analysis includes many more `r TAG`. In order to include the not-annotated `r TAG` in the
enrichment analysis, to see if annotated or not annotated `r TAG` are more or less often
differentially expressed, I should attribute a GO term to them. According to [http://geneontology.org/docs/faq/]
nowadays we express lack of annotation by annotating to the root nodes, i.e. GO:0008150 biological_process,
GO:0003674 molecular_function, and GO:0005575 cellular_component.

```{r unknown}
for (tag in names(TStats)) {
   if (! tag %in% names(allgenes2GO)) {
      allgenes2GO <- append(allgenes2GO,
         structure(list(c("GO:0008150", "GO:0003674", "GO:0005575")), names = tag))
   }
}
```

# Building the topGO object

Creation of a topGO dataset is documented in section 4 of topGO's the user manual:
https://bioconductor.org/packages/release/bioc/vignettes/topGO/inst/doc/topGO.pdf.
I need to use the `new`
command, and fill up the slots. The `annot` object must be a function that compiles "a list of
GO terms such that each element in the list is a character vector containing all the gene 
identifiers that are mapped to the respective GO term." There are several options, that you can
check using `help(annFUN.gene2GO)`, for example. The `annFUN.gene2GO` requires a `gene2GO`
argument, which is the list of gene-to-GO terms I made before. The `geneSelectionFun` object is a
function that selects the *interesting* (most significant) genes. It is required to
perform Fisher's exact test. The `nodeSize` is used to prune the GO hierarchy from the terms
which have less than `n` annotated genes.

I generate three datasets, to analyse each of the three ontologies.

```{r GOdata1, results = 'hold', message = FALSE}
selection <- function(allScores) {return(abs(allScores) > 5.0)}
GOdata.BP <- new('topGOdata',
   description = 'Differential expression among Brachionus plicatilis populations.',
   ontology = 'BP', 
   allGenes = TStats,
   annot = annFUN.gene2GO,
   gene2GO = allgenes2GO,
   geneSelectionFun = selection,
   nodeSize = 5)

GOdata.MF <- new('topGOdata',
   description = 'Differential expression among Brachionus plicatilis populations.',
   ontology = 'MF',
   allGenes = TStats,
   annot = annFUN.gene2GO,
   gene2GO = allgenes2GO,
   geneSelectionFun = selection,
   nodeSize = 5)

GOdata.CC <- new('topGOdata',
   description = 'Differential expression among Brachionus plicatilis populations.',
   ontology = 'CC',
   allGenes = TStats,
   annot = annFUN.gene2GO,
   gene2GO = allgenes2GO,
   geneSelectionFun = selection,
   nodeSize = 5)

DataSummary <- data.frame(ontology = c('BP', 'MF', 'CC'),
   Num_Genes = sapply(list(GOdata.BP, GOdata.MF, GOdata.CC), numGenes),
   Num_GO_terms = sapply(list(GOdata.BP, GOdata.MF, GOdata.CC), function(x) length(usedGO(x))))
kable(DataSummary, caption='Number of feasible genes or transcripts and number of GO terms used in each data set.')
rm(DataSummary)
```

# Running the tests

There are more than one way to test for enrichment. Something that took me a while to understand
is that not only there are different test statistics (Fisher's exact test, Kolmogorov-Smirnov,
and others) but also different *algorithms*: classic, elim, weight... The algorithms are ways
to deal with the dependence structure among GO terms due to topology. Some algorithms are
compatible with all statistics implemented in topGO. But the *weight* and the *parentchild*
algorithms are only applicable to Fisher's exact test. I am not interested in the classic
algorithm, which treats GO nodes as independent, and therefore produces an excess of significant
results. I will not use the Fisher's exact test, because it dependes on an arbitrary threshold
of significance on non-adjusted p-values.

```{r testing1, message = FALSE}
BP.elim     <- runTest(GOdata.BP, algorithm = "elim",     statistic = "ks")
BP.weight01 <- runTest(GOdata.BP, algorithm = "weight01", statistic = "ks")
BP.lea      <- runTest(GOdata.BP, algorithm = "lea",      statistic = "ks")
MF.elim     <- runTest(GOdata.MF, algorithm = "elim",     statistic = "ks")
MF.weight01 <- runTest(GOdata.MF, algorithm = "weight01", statistic = "ks")
MF.lea      <- runTest(GOdata.MF, algorithm = "lea",      statistic = "ks")
CC.elim     <- runTest(GOdata.CC, algorithm = "elim",     statistic = "ks")
CC.weight01 <- runTest(GOdata.CC, algorithm = "weight01", statistic = "ks")
CC.lea      <- runTest(GOdata.CC, algorithm = "lea",      statistic = "ks")

ResultsSummary <- data.frame(ontology = rep(c("BP", "MF", "CC"), each = 3),
   algorithm = rep(c("elim", "weight01", "lea"), 3),
   TermsTested = sapply(list(BP.elim, BP.weight01, BP.lea, MF.elim, MF.weight01, MF.lea, CC.elim, CC.weight01, CC.lea), function(X) length(score(X))),
   Significant = sapply(list(BP.elim, BP.weight01, BP.lea, MF.elim, MF.weight01, MF.lea, CC.elim, CC.weight01, CC.lea), function(X) sum(score(X) < 0.01)))

kable(ResultsSummary, caption="Number of non-trivial terms tested and those with a score (not corrected p-value) lower than 0.01.")
rm(ResultsSummary)
```

# Results

I don't see a way to distinguish GO terms that are significant because they tend to
appear near the top of the list of genes from those that are significant because they
tend to appear near the bottom. The density plots are useful, but take too much space
to show them for all terms.

Note that not all terms listed in the tables below are significant. The scores for
the three methods (elim, weight01 and lea) are non-corrected p-values.

## Biological process

```{r BP1, message=FALSE}
orderedTerms <- names(sort(score(BP.weight01)))
significant.weight01 <- score(BP.weight01)[orderedTerms] <= 0.01
significant.lea      <- score(BP.lea)[orderedTerms] <= 0.01
significant.elim     <- score(BP.elim)[orderedTerms] <= 0.01
sigTerms <- orderedTerms[significant.weight01 & significant.lea & significant.elim]
# sigTerms gets updated, and is already used in the code, but I need a permanent copy for later.
BP.tstat.sigTerms <- sigTerms

BP.all <- GenTable(GOdata.BP, elim=BP.elim, weight01=BP.weight01, lea=BP.lea,
                   orderBy="weight01", ranksOf="elim", topNodes=sum(significant.elim))

kable(
   BP.all[BP.all$Significant > BP.all$Expected,],
   caption = "Most over-represented terms of the Biological Process ontology.")

kable(
   BP.all[BP.all$Significant < BP.all$Expected,],
   caption = "Most under-represented terms of the Biological Process ontology.")

vennDiagram(vennCounts(cbind(weight01=significant.weight01,
                             lea=significant.lea,
                             elim=significant.elim)))

kable(data.frame(Term = Term(GOTERM[sigTerms]),
                 Definition = Definition(GOTERM[sigTerms]),
                 PValue=score(BP.weight01)[sigTerms]),
      caption = paste('Biological process terms significantly associated with',
                      VAR, 'according to all 3 algorithms', sep=' '))
```

I think the GO graph is useful to see the relationship among the significant terms. But
too large graphs are impossible to read. I don't know how to split the graph in meaningful
subgraphs.

```{r BP1dag, message=FALSE, fig.width=30, fig.height=15}
showSigOfNodes(GOdata.BP, score(BP.weight01),
               firstSigNodes = sum(significant.elim),
               wantedNodes = sigTerms)
```

This is just a example of the most significant GO term:

```{r BP1density, message=FALSE}
showGroupDensity(GOdata.BP, orderedTerms[1], rm.one=FALSE)
showGroupDensity(GOdata.BP, orderedTerms[2], rm.one=FALSE)
showGroupDensity(GOdata.BP, orderedTerms[3], rm.one=FALSE)
```

## Molecular function

```{r MF1, message=FALSE}
orderedTerms <- names(sort(score(MF.weight01)))
significant.weight01 <- score(MF.weight01)[orderedTerms] <= 0.01
significant.lea      <- score(MF.lea)[orderedTerms] <= 0.01
significant.elim     <- score(MF.elim)[orderedTerms] <= 0.01
sigTerms <- orderedTerms[significant.weight01 & significant.lea & significant.elim]
# sigTerms gets updated, and is already used in the code, but I need a permanent copy for later.
MF.tstat.sigTerms <- sigTerms

MF.all <- GenTable(GOdata.MF, elim=MF.elim, weight01=MF.weight01, lea=MF.lea,
                   orderBy="weight01", ranksOf="elim", topNodes=sum(significant.elim))

kable(
   MF.all[MF.all$Significant > MF.all$Expected,],
   caption = "Most over-represented terms of the Molecular Function ontology.")

kable(
   MF.all[MF.all$Significant < MF.all$Expected,],
   caption = "Most under-represented terms of the Molecular Function ontology.")

vennDiagram(vennCounts(cbind(weight01=significant.weight01,
                             lea=significant.lea,
                             elim=significant.elim)))

kable(data.frame(Term=Term(GOTERM[sigTerms]),
                 Definition=Definition(GOTERM[sigTerms]),
                 PValue=score(MF.weight01)[sigTerms]),
      caption = paste('Molecular function terms significantly associated with', VAR,
                      'according to all 3 algorithms', sep=' '))
```

```{r MF1dag, message=FALSE, fig.width=30, fig.height=15}
showSigOfNodes(GOdata.MF, score(MF.weight01),
               firstSigNodes = sum(significant.elim),
               wantedNodes = sigTerms)
```

```{r MF1density, message=FALSE}
showGroupDensity(GOdata.MF, orderedTerms[1], rm.one=FALSE)
showGroupDensity(GOdata.MF, orderedTerms[2], rm.one=FALSE)
showGroupDensity(GOdata.MF, orderedTerms[3], rm.one=FALSE)
```

## Cellular component

```{r CC1, message=FALSE}
orderedTerms <- names(sort(score(CC.weight01)))
significant.weight01 <- score(CC.weight01)[orderedTerms] <= 0.01
significant.lea      <- score(CC.lea)[orderedTerms] <= 0.01
significant.elim     <- score(CC.elim)[orderedTerms] <= 0.01
sigTerms <- orderedTerms[significant.weight01 & significant.lea & significant.elim]
# sigTerms gets updated, and is already used in the code, but I need a permanent copy for later.
CC.tstat.sigTerms <- sigTerms

CC.all <- GenTable(GOdata.CC, elim=CC.elim, weight01=CC.weight01, lea=CC.lea,
                   orderBy="weight01", ranksOf="elim", topNodes=sum(significant.elim))
kable(
   CC.all[CC.all$Significant > CC.all$Expected,],
   caption = "Most over-represented terms of the Cellular Component ontology.")

kable(
   CC.all[CC.all$Significant < CC.all$Expected,],
   caption = "Most under-represented terms of the Cellular Component ontology.")

vennDiagram(vennCounts(cbind(weight01=significant.weight01,
                             lea=significant.lea,
                             elim=significant.elim)))

kable(data.frame(Term=Term(GOTERM[sigTerms]),
                 Definition=Definition(GOTERM[sigTerms]),
                 PValue=score(CC.weight01)[sigTerms]),
      caption = paste('Cellular component terms significantly associated with', VAR,
                      'according to all 3 algorithms', sep=' '))
```

```{r CC1dag, message=FALSE, fig.width=30, fig.height=15}
showSigOfNodes(GOdata.CC, score(CC.weight01),
               firstSigNodes = sum(significant.elim),
               wantedNodes = sigTerms)
```

```{r CC1density, message=FALSE}
showGroupDensity(GOdata.CC, orderedTerms[1], rm.one=FALSE)
showGroupDensity(GOdata.CC, orderedTerms[2], rm.one=FALSE)
showGroupDensity(GOdata.CC, orderedTerms[3], rm.one=FALSE)
```

# Comparison between the two ordering of genes

I want to compare these results with those from `2020-01-14`, where I used
the $p$ value of the differential expression analysis to order the genes.
I can import the data from `2020-01-14`, but I need to do it in a new
environment to prevent overwriting the current results.

```{r newEnv}
load(ENRICHMENT, ex <- new.env())
```

## Biological process

```{r BPcomparison, message=FALSE}
allTerms <- usedGO(GOdata.BP)
BP.pvalue.sigTerms <- with(ex, BP.pvalue.sigTerms)
vennDiagram(vennCounts(cbind(TStat  = allTerms %in% BP.tstat.sigTerms,
                             PValue = allTerms %in% BP.pvalue.sigTerms)))
# New terms:
as.data.frame(Term(GOTERM[setdiff(BP.tstat.sigTerms, BP.pvalue.sigTerms)]))

# Absent terms:
as.data.frame(Term(GOTERM[setdiff(BP.pvalue.sigTerms, BP.tstat.sigTerms)]))

BPp.weight01 <- with(ex, BP.weight01)
ggplot(data = data.frame(TStat  = rank(score(BP.weight01))[allTerms],
                         PValue = rank(score(BPp.weight01))[allTerms]),
       mapping = aes(x = TStat, y = PValue)) +
   geom_point() + geom_smooth() + xlab('Using gene t statistics') +
   ylab('Using gene p values') +
   ggtitle('Ordering of BP terms by significance')
```

## Molecular function

```{r MFcomparison, message=FALSE}
allTerms <- usedGO(GOdata.MF)
MF.pvalue.sigTerms <- with(ex, MF.pvalue.sigTerms)
vennDiagram(vennCounts(cbind(TStat  = allTerms %in% MF.tstat.sigTerms,
                             PValue = allTerms %in% MF.pvalue.sigTerms)))
# New terms:
as.data.frame(Term(GOTERM[setdiff(MF.tstat.sigTerms, MF.pvalue.sigTerms)]))

# Absent terms:
as.data.frame(Term(GOTERM[setdiff(MF.pvalue.sigTerms, MF.tstat.sigTerms)]))

MFp.weight01 <- with(ex, MF.weight01)
ggplot(data = data.frame(TStat  = rank(score(MF.weight01))[allTerms],
                         PValue = rank(score(MFp.weight01))[allTerms]),
       mapping = aes(x = TStat, y = PValue)) +
   geom_point() + geom_smooth() + xlab('Using gene t statistics') +
   ylab('Using gene p values') +
   ggtitle('Ordering of MF terms by significance')
```

## Cellular component

```{r CCcomparison, message=FALSE}
allTerms <- usedGO(GOdata.CC)
CC.pvalue.sigTerms <- with(ex, CC.pvalue.sigTerms)
vennDiagram(vennCounts(cbind(TStat  = allTerms %in% CC.tstat.sigTerms,
                             PValue = allTerms %in% CC.pvalue.sigTerms)))
# New terms:
as.data.frame(Term(GOTERM[setdiff(CC.tstat.sigTerms, CC.pvalue.sigTerms)]))

# Absent terms:
as.data.frame(Term(GOTERM[setdiff(CC.pvalue.sigTerms, CC.tstat.sigTerms)]))

CCp.weight01 <- with(ex, CC.weight01)
ggplot(data = data.frame(TStat  = rank(score(CC.weight01))[allTerms],
                         PValue = rank(score(CCp.weight01))[allTerms]),
       mapping = aes(x = TStat, y = PValue)) +
   geom_point() + geom_smooth() + xlab('Using gene t statistics') +
   ylab('Using gene p values') +
   ggtitle('Ordering of CC terms by significance')
```

As expected, ordering genes by the $t$ statistic produces results different from
those when ordering genes by $p$ value. Some GO terms enriched when using $p$ values
are missed when using $t$ statistic, typically because those terms are assigned
to genes both sub- and overexpressed. See for example the case of the protein
phosphorylation process:

```{r snake}
showGroupDensity(GOdata.BP, 'GO:0006468', rm.one=FALSE)
```

Both the genes annotated with that function and their complement display a
snake-shaped, bimodal distribution, the former being more extreme. It is not
irrelevant, but more difficult to interprete than a term being significant
because all their genes behave in a similar way.

# Session info

```{r last}
save(allgenes2GO,
     GOdata.BP, BP.elim, BP.weight01, BP.lea, BP.pvalue.sigTerms,
     GOdata.MF, MF.elim, MF.weight01, MF.lea, MF.pvalue.sigTerms,
     GOdata.CC, CC.elim, CC.weight01, CC.lea, CC.pvalue.sigTerms,
     file = paste('Enrichment', TAG, VAR, 'RData', sep='.'))
sessionInfo()
```