---
title: "Population-specific effects of randomness on gene expression"
author: "J. Ignacio Lucas Lledó"
date: "23/12/2019"
output: html_document
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(edgeR)
library(ggplot2)
library(reactable)
library(cp4p)
COUNTS_FILE='../2019-03-29/genes.PostCount.txt'
```

One of the conclusions in `2019-04-03` was that there is a significant interaction between the
effects of the regime and of the original population. That means that not all populations responded
equally to a specific selective regime. We need to take this into account to decide what contrasts
are interesting. Not only the genes that are *on average* differently expressed in regular than
in random environments are interesting. The specific response of each random population may have
biological meaning.

On `2019-04-03` we also determined that there are no genes with a significant interaction between
selective regime and hatching condition. This may be a simple consequence of the fact that not
all populations respond similarly to the selective regimes: in this situation, there are just
not enough replicates to detect an interaction between regime and hatching condition.

# Genes expressed *on average* differently between random and regular regimes

When I first tested the effect of randomness on expression level using model 4 and 6 on `2019-04-03`,
I tested for a significant second coefficient, which corresponds to the expression level of a
sample from original population 1 under a random regime, relative to the expression level of population
3, under a regular regime. That does not seem to be what I want. To make it easier to express a contrast
that compares all random populations to all regular ones, I will use a different model.

```{r variables, cache=TRUE}
counts <- read.table(COUNTS_FILE, row.names=1, header=TRUE)
regimePop <- factor(c('random1',  'random1',  'random2',  'random2',
                      'regular1', 'regular1', 'random3',  'random3',
                      'regular2', 'regular2', 'regular3', 'regular3'))
diapause   <- factor(rep(c("hatch","diapause"), 6))
diapause   <- relevel(diapause, "hatch")

# Filtering
y <- DGEList(counts=counts, group=regimePop)
y <- calcNormFactors(y, method="TMM")
threshold <- 5.0 / (min(y$samples$lib.size) / 1000000)
keep <- rowSums(cpm(y) > threshold) >= 4
y <- y[keep, ,keep.lib.sizes=FALSE]
NumOfTags <- dim(y)[1]
design7 <- model.matrix(~ 0 + regimePop + diapause)
row.names(design7) <- names(counts)
reactable(design7, columns=list(
  regimePoprandom1 = colDef(name='Random.1'),
  regimePoprandom2 = colDef(name='Random.2'),
  regimePoprandom3 = colDef(name='Random.3'),
  regimePopregular1 = colDef(name='Regular.1'),
  regimePopregular2 = colDef(name='Regular.2'),
  regimePopregular3 = colDef(name='Regular.3'),
  diapausediapause = colDef(name='diapause')))
y7 <- estimateDisp(y, design7)
fit7 <- glmQLFit(y7, design7)
randomness <- glmQLFTest(fit7, contrast = c(1/3, 1/3, 1/3, -1/3, -1/3, -1/3, 0))
```

By treating populations independently, model 7 neither assumes nor estimates whatever common effect
populations in the same regime may share.

Instead of reporting the number of significantly regulated genes under a specific FDR,
I report here the overall, FDR-independent, estimated fraction of regulated
genes. There are ways to estimate the number of true null hypotheses even when
there is a dependence structure among tests. I found the package `cp4p` (Gianetto et al.
2016. Proteomics, 16(1), 29-32.), that does exactly that. The SLIM method [Want et al. 2011](https://academic.oup.com/bioinformatics/article/27/2/225/286449),
is supposed to be adequate for datasets with dependence structure, but I think here it
overestimates the proportion of non-regulated genes.


```{r randomness, cache=TRUE}
randomness.pi0 <- estim.pi0(randomness$table$PValue)
randomness.pi0
calibration.plot(randomness$table$PValue, pi0.method = 'langaas')
topTags(randomness)
z1 <- topTags(randomness, n=NumOfTags)$table
z1$logPValue <- -log10(z1$PValue)
ggplot(data=z1, mapping=aes(x=logFC, y=logPValue, color=FDR)) + 
   geom_point() + theme_grey() + ylab(expression(paste("-log"[10], plain("(p value)")))) +
   ggtitle('Random versus regular. Model 7.')
rm(z1)
```

According to this test, at least `r round(1 - max(randomness.pi0$pi0.est), 2) * 100` and
up to `r round(1.0 - min(randomness.pi0$pi0.est), 2) * 100`% of genes are differentially expressed
on average between regular and random regimes.

# Genes expressed differently in random population 1 than in the average regular population

Using again model 7, now I want to determine what genes are expressed differently in random
population 1, with respect to the average regular population.

```{r random1, cache=TRUE}
random1 <- glmQLFTest(fit7, contrast = c(1, 0, 0, -1/3, -1/3, -1/3, 0))
random1.pi0 <- estim.pi0(random1$table$PValue)
random1.pi0
calibration.plot(random1$table$PValue, pi0.method = 'langaas')
topTags(random1)
z1 <- topTags(random1, n=NumOfTags)$table
z1$logPValue <- -log10(z1$PValue)
ggplot(data=z1, mapping=aes(x=logFC, y=logPValue, color=FDR)) + 
   geom_point() + theme_grey() + ylab(expression(paste("-log"[10], plain("(p value)")))) +
   ggtitle('Random:1 versus regular. Model 7.')
rm(z1)
```

According to this test, at least `r round(1 - max(random1.pi0$pi0.est), 2) * 100` and
up to `r round(1.0 - min(random1.pi0$pi0.est), 2) * 100`% of genes are expressed differently
in population 1, under a random regime, relative to the average across populations in the
regular regime.

# Genes expressed differently in random population 2 than in the average regular population

Next, we examine population 2, also under a random regime, relative to the average of populations
under the regular regime.

```{r random2, cache=TRUE}
random2 <- glmQLFTest(fit7, contrast = c(0, 1, 0, -1/3, -1/3, -1/3, 0))
random2.pi0 <- estim.pi0(random2$table$PValue)
random2.pi0
calibration.plot(random2$table$PValue, pi0.method = 'langaas')
topTags(random2)
z1 <- topTags(random2, n=NumOfTags)$table
z1$logPValue <- -log10(z1$PValue)
ggplot(data=z1, mapping=aes(x=logFC, y=logPValue, color=FDR)) + 
   geom_point() + theme_grey() + ylab(expression(paste("-log"[10], plain("(p value)")))) +
   ggtitle('Random:2 versus regular. Model 7.')
rm(z1)
```

According to this test, at least `r round(1 - max(random2.pi0$pi0.est), 2) * 100` and
up to `r round(1.0 - min(random2.pi0$pi0.est), 2) * 100`% of genes are expressed differently
in population 2, under a random regime, relative to the average across populations in the
regular regime.

# Genes expressed differently in (random) population 4 than in the average regular population

The third population under a random regime is labeled *4* in samples names. Let's see what and
how many genes are expressed differently in that population, relative to those under regular
regime:

```{r random3, cache=TRUE}
random3 <- glmQLFTest(fit7, contrast = c(0, 0, 1, -1/3, -1/3, -1/3, 0))
random3.pi0 <- estim.pi0(random3$table$PValue)
random3.pi0
calibration.plot(random3$table$PValue, pi0.method = 'langaas')
topTags(random3)
z1 <- topTags(random3, n=NumOfTags)$table
z1$logPValue <- -log10(z1$PValue)
ggplot(data=z1, mapping=aes(x=logFC, y=logPValue, color=FDR)) + 
   geom_point() + theme_grey() + ylab(expression(paste("-log"[10], plain("(p value)")))) +
   ggtitle('Random:3 versus regular. Model 7.')
rm(z1)
```

According to this test, at least `r round(1 - max(random3.pi0$pi0.est), 2) * 100` and
up to `r round(1.0 - min(random3.pi0$pi0.est), 2) * 100`% of genes are expressed differently
in population 3, under a random regime, relative to the average across populations in the
regular regime.


# Genes always expressed differently in a random population than in the average regular population

The scatterplot of p-values is not expected to produce any visible correlation, unless for the smallest
p-values. Plotting p-values in logarithmic scale does show that many genes differentially expressed
in any population under the random regime is also differentially expressed in the other populations
in the same regime.

```{r scatterplot}
pairs(data.frame(rand1=log(random1$table$PValue),
                 rand2=log(random2$table$PValue),
                 rand3=log(random3$table$PValue)),
      pch=".", main='Log(p-values)')
```

What about the log-fold change? Does the random regime change gene expression in the same direction in
all populations? Yes:

```{r logFoldChange}
pairs(data.frame(rand1=random1$table$logFC,
                 rand2=random2$table$logFC,
                 rand3=random3$table$logFC),
      pch = ".", main = 'Log(fold change)')
```

Given the large number of regulated genes, I can use a conservative FDR to compare the most significantly
regulated genes among populations in the random regime.

```{r venn}
significant <- cbind(randomness = topTags(randomness, n=NumOfTags, sort.by='none', adjust.method='fdr')$table$FDR <= 0.1, 
                     random1 = topTags(random1, n=NumOfTags, sort.by='none', adjust.method='fdr')$table$FDR <= 0.1,
                     random2 = topTags(random2, n=NumOfTags, sort.by='none', adjust.method='fdr')$table$FDR <= 0.1,
                     random3 = topTags(random3, n=NumOfTags, sort.by='none', adjust.method='fdr')$table$FDR <= 0.1)
vennDiagram(vennCounts(significant), main='FDR=0.1')
```

# Writing up the results

A large proportion of genes up- or downregulated in any population under the random regime is also regulated
in at least one other population under the random regime. The first and third populations in the random regime
are more similar among them. Population 2 being the one with more genes regulated by the selective regime is also
the one with the largest proportion of exclusively regulated genes.

I should save the results of the `glmQLFTest()` to run the functional analysis on the lists of regulated genes in
a different session. I think it's also good practice to save a plain text copy of the main tables, to allow
inspection outside R.

```{r saving}
save(randomness, random1, random2, random3, file = 'genes/RegimeEffect_model7.RData')
write.table(topTags(randomness, n=NumOfTags)[[1]],
            file = 'genes/randomness.txt',
            quote = FALSE)
write.table(topTags(random1, n=NumOfTags)[[1]],
            file = 'genes/random1.txt',
            quote = FALSE)
write.table(topTags(random2, n=NumOfTags)[[1]],
            file = 'genes/random2.txt',
            quote = FALSE)
write.table(topTags(random3, n=NumOfTags)[[1]],
            file = 'genes/random3.txt',
            quote = FALSE)
```
