---
title: "Functional Enrichment Analysis"
output: html_document
author: "J. Ignacio Lucas Lledó"
---

```{r eval=TRUE, echo=FALSE, include=FALSE}
library(topGO)
```

# Reading the data
In folder 2019-04-03 I produced the lists of genes and isoforms that underwent the differential
expression analysis, with the fold change, the p-values, etc. In addition, some transcripts were
annotated in folder 2019-07-19 with Gene Ontology terms. From that analysis, I produced here the
lists of annotated genes and transcripts with their attributed GO terms. TopGO requires both the
genes (or transcripts) with an ordering criteria (e.g., p-value), and the annotation, which can
be uploaded as a named list.

Here, I want to re-use this script to run the enrichment analysis of both genes and isoforms, and
both for gene expression differences in between selective regimes and between hatching conditions.
Let's read the arguments to figure what what I am doing now.

It was not trivial to get `grep` below return a single logical value. I need to use `any(grep(...))`.

```{r}
args <- commandArgs(trailingOnly = TRUE)
if (length(args) < 2) {
   stop("I need the file names of the list of genes or isoforms and their annotations.", call.=FALSE)
} else {
   list_file <- args[1]
   annotation_file <- args[2]
}

if (any(grep("gene", annotation_file, fixed=TRUE))) {
   TagType = "genes"
} else {
   if (any(grep("trans", annotation_file, fixed=TRUE))) {
      TagType = "transcripts"
   } else {
      TagType = "unknown tags!"
   }
}

if (any(grep("regime", list_file, fixed=TRUE))) {
   AnalysisType = "selective regimes"
} else {
   if (any(grep("hatching", list_file, fixed=TRUE))) {
      AnalysisType = "hatching conditions"
   } else {
     AnalysisType = "unknown conditions"
   }
}
```

So, here I am running a functional enrichment analysis among `r TagType` differentially expressed
between `r AnalysisType`. The `list_file`, here `r list_file`, includes all the `r TagType` that
passed the filters, and is already formated for the header to be recognized. I only need a named
list of p-values, where the names correspond to the `r TagType`. Thus, I skip the columns containing
the log fold change, the log count per million, the F statistic and the false discovery rate.

```{r}
tags <- read.table(list_file, colClasses = c("character", "NULL", "NULL", "NULL", "numeric", "NULL"))
PValues <- structure(tags$PValue, names = row.names(tags))
rm(tags)
```

There are `r length(PValues)` `r TagType`. Now, I load the functional annotation of the `r TagType`
as a named list. 

```{r}
annotation <- read.table(annotation_file, col.names = c('tagname', 'goterms'))
allgenes2GO <- strsplit(as.character(annotation$goterms), "|", fixed = TRUE)
names(allgenes2GO) <- annotation$tagname
rm(annotation)
```

There are `r length(allgenes2GO)` `r TagType` with GO annotations. But the differential expression
analysis includes many more `r TagType`. If I wanted to include the not-annotated `r TagType` in the
enrichment analysis, to see if annotated or not annotated `r TagType` are more or less often
differentially expressed, I should attribute a GO term to them. According to http://geneontology.org/docs/faq/
nowadays we express lack of annotation by annotating to the root nodes, i.e. GO:0008150 biological_process,
GO:0003674 molecular_function, and GO:0005575 cellular_component.

```{r}
for (tag in names(PValues)) {
   if (! tag %in% names(allgenes2GO)) {
      allgenes2GO <- append(allgenes2GO,
         structure(list(c("GO:0008150", "GO:0003674", "GO:0005575")), names = tag))
   }
}
```

# Building the topGO object

Creation of a topGO dataset is documented in section 4 of topGO's the user manual:
https://bioconductor.org/packages/release/bioc/vignettes/topGO/inst/doc/topGO.pdf.
I need to use the `new`
command, and fill up the slots. The `annot` object must be a function that compiles "a list of
GO terms such that each element in the list is a character vector containing all the gene 
identifiers that are mapped to the respective GO term." There are several options, that you can
check using `help(annFUN.gene2GO)`, for example. The `annFUN.gene2GO` requires a `gene2GO`
argument, which is the list of gene-to-GO terms I made before. The `geneSelectionFun` object is a
function that selects the *interesting* (most significant) genes. It is required to
perform Fisher's exact test. The `nodeSize` is used to prune the GO hierarchy from the terms
which have less than `n` annotated genes.

I generate three datasets, to analyse each of the three ontologies.

```{r results = 'hold', message = FALSE}
selection <- function(allScores) {return(allScores < 0.05)}
GOdata.BP <- new('topGOdata',
   description = 'Differential expression among Brachionus plicatilis populations.',
   ontology = 'BP', 
   allGenes = PValues,
   annot = annFUN.gene2GO,
   gene2GO = allgenes2GO,
   geneSelectionFun = selection,
   nodeSize = 5)

GOdata.MF <- new('topGOdata',
   description = 'Differential expression among Brachionus plicatilis populations.',
   ontology = 'MF',
   allGenes = PValues,
   annot = annFUN.gene2GO,
   gene2GO = allgenes2GO,
   geneSelectionFun = selection,
   nodeSize = 5)

GOdata.CC <- new('topGOdata',
   description = 'Differential expression among Brachionus plicatilis populations.',
   ontology = 'CC',
   allGenes = PValues,
   annot = annFUN.gene2GO,
   gene2GO = allgenes2GO,
   geneSelectionFun = selection,
   nodeSize = 5)

library(knitr)
DataSummary <- data.frame(ontology = c('BP', 'MF', 'CC'),
   Num_Genes = sapply(list(GOdata.BP, GOdata.MF, GOdata.CC), numGenes),
   Num_GO_terms = sapply(list(GOdata.BP, GOdata.MF, GOdata.CC), function(x) length(usedGO(x))))
kable(DataSummary, caption='Number of feasible genes or transcripts and number of GO terms used in each data set.')
rm(DataSummary)
```

# Running the tests

There are more than one way to test for enrichment. Something that took me a while to understand
is that not only there are different test statistics (Fisher's exact test, Kolmogorov-Smirnov,
and others) but also different *algorithms*: classic, elim, weight... The algorithms are ways
to deal with the dependence structure among GO terms due to topology. Some algorithms are
compatible with all statistics implemented in topGO. But the *weight* and the *parentchild*
algorithms are only applicable to Fisher's exact test. I am not interested in the classic
algorithm, which treats GO nodes as independent, and therefore produces an excess of significant
results. I will not use the Fisher's exact test, because it dependes on an arbitrary threshold
of significance on non-adjusted p-values.

```{r results = 'asis', message = FALSE}
BP.elim     <- runTest(GOdata.BP, algorithm = "elim",     statistic = "ks")
BP.weight01 <- runTest(GOdata.BP, algorithm = "weight01", statistic = "ks")
BP.lea      <- runTest(GOdata.BP, algorithm = "lea",      statistic = "ks")
MF.elim     <- runTest(GOdata.MF, algorithm = "elim",     statistic = "ks")
MF.weight01 <- runTest(GOdata.MF, algorithm = "weight01", statistic = "ks")
MF.lea      <- runTest(GOdata.MF, algorithm = "lea",      statistic = "ks")
CC.elim     <- runTest(GOdata.CC, algorithm = "elim",     statistic = "ks")
CC.weight01 <- runTest(GOdata.CC, algorithm = "weight01", statistic = "ks")
CC.lea      <- runTest(GOdata.CC, algorithm = "lea",      statistic = "ks")

ResultsSummary <- data.frame(ontology = rep(c("BP", "MF", "CC"), each = 3),
   algorithm = rep(c("elim", "weight01", "lea"), 3),
   TermsTested = sapply(list(BP.elim, BP.weight01, BP.lea, MF.elim, MF.weight01, MF.lea, CC.elim, CC.weight01, CC.lea), function(X) length(score(X))),
   Significant = sapply(list(BP.elim, BP.weight01, BP.lea, MF.elim, MF.weight01, MF.lea, CC.elim, CC.weight01, CC.lea), function(X) sum(score(X) < 0.01)))

kable(ResultsSummary, caption="Number of non-trivial terms tested and those with a score (not corrected p-value) lower than 0.01.")
rm(ResultsSummary)
```

# Results

Curiously, topGO does not pay much attention to what terms are significant because they are
_over_represented and which ones are _under_represented. I think it's worth separating them,
to facilitate the biological interpretation. Note that not all terms listed in the tables
below are significant. The scores for the three methods (elim, weight01 and lea) are non-corrected
p-values.

```{r message = FALSE, results='asis'}
BP.all <- GenTable(GOdata.BP, elim=BP.elim, weight01=BP.weight01, lea=BP.lea, orderBy="elim", ranksOf="weight01", topNodes=50)
MF.all <- GenTable(GOdata.MF, elim=MF.elim, weight01=MF.weight01, lea=MF.lea, orderBy="elim", ranksOf="weight01", topNodes=50)
CC.all <- GenTable(GOdata.CC, elim=CC.elim, weight01=CC.weight01, lea=CC.lea, orderBy="elim", ranksOf="weight01", topNodes=50)

kable(
   BP.all[BP.all$Significant > BP.all$Expected,],
   caption = "Most over-represented terms of the Biological Process ontology.")

kable(
   BP.all[BP.all$Significant < BP.all$Expected,],
   caption = "Most under-represented terms of the Biological Process ontology.")

kable(
   MF.all[MF.all$Significant > MF.all$Expected,],
   caption = "Most over-represented terms of the Molecular Function ontology.")

kable(
   MF.all[MF.all$Significant < MF.all$Expected,],
   caption = "Most under-represented terms of the Molecular Function ontology.")

kable(
   CC.all[CC.all$Significant > CC.all$Expected,],
   caption = "Most over-represented terms of the Cellular Component ontology.")

kable(
   CC.all[CC.all$Significant < CC.all$Expected,],
   caption = "Most under-represented terms of the Cellular Component ontology.")
```
