---
title: "Functional enrichment analysis"
output: html_document
author: "J. Ignacio Lucas Lledó"
params:
   TAG: 'genes'
   VAR: 'regime'
---

```{r setup, message=FALSE}
library(topGO)
library(knitr)   # to use kable()
library(limma)   # to use vennDiagram()
library(ggplot2)
TAG <- params$TAG
VAR <- params$VAR
ANNOTATION <- list(genes = '../2019-07-26/genes/annotation.txt',
                   isoforms = '../2019-07-26/transcripts/annotation.txt')
TAGDIR <- paste0('../2020-01-08/', TAG, '/')
```

# Introduction
This is the enrichment analysis for `r TAG` regulated by `r VAR`. Because I quantified the association of expression
with `r VAR` in two different ways, this document has two sections. Section *Variance* reports the enrichment
analysis performed with `r TAG` ordered by the proportion of expression variance explained by `r VAR`. Note that
some `r TAG` with low variance may still be highly associated with `r VAR`, even if the fold change between
levels of this factor is low. Section *Differential expression* uses an ordering of `r TAG` based on the significance
of the differential expression between levels of `r VAR`, which does depend on fold change.

# Reading the data
Functional annotation is in `2019-07-26`. I will also upload two lists of `r TAG`, with either proportion of
variance explained by `r VAR` or p-value of differential expression test.

```{r data}
tagVariance <- read.table(paste0(TAGDIR, VAR, '_variance.txt'))
tagPValue   <- read.table(paste0(TAGDIR, VAR, '_pvalue.txt'))
annotation  <- read.table(ANNOTATION[[TAG]], col.names = c('tagname', 'goterms'))
```

To initialize the `topGOdata` object, I will need the gene list as a named numeric vector, where the
names are the `r TAG` identifiers and the numeric values, either the portion of variance explained by
`r VAR` or the p-values of the differential expression test. The `structure()` function adds attributes
to an object.

```{r scores}
Variance <- structure(tagVariance[,1], names = row.names(tagVariance))
PValues  <- structure(tagPValue[,1],   names = row.names(tagPValue))
rm(tagVariance, tagPValue)
```

There are `r length(Variance)` `r TAG` scored with a variance portion and a differential expression p-value.
It should actually be the exact same `r TAG`.
The `annotation` data frame has more than one GO term for every tag, separated by `|`. I need a named list.

```{r annotation}
head(annotation)
allgenes2GO <- strsplit(as.character(annotation$goterms), "|", fixed = TRUE)
names(allgenes2GO) <- annotation$tagname
rm(annotation)
```

There are `r length(allgenes2GO)` `r TAG` with GO annotations. But the differential expression
analysis includes many more `r TAG`. In order to include the not-annotated `r TAG` in the
enrichment analysis, to see if annotated or not annotated `r TAG` are more or less often
differentially expressed, I should attribute a GO term to them. According to [http://geneontology.org/docs/faq/]
nowadays we express lack of annotation by annotating to the root nodes, i.e. GO:0008150 biological_process,
GO:0003674 molecular_function, and GO:0005575 cellular_component.

```{r unknown}
for (tag in unique(c(names(PValues), names(Variance)))) {
   if (! tag %in% names(allgenes2GO)) {
      allgenes2GO <- append(allgenes2GO,
         structure(list(c("GO:0008150", "GO:0003674", "GO:0005575")), names = tag))
   }
}
```

# Using differential expression p-values
## Building the topGO object

Creation of a topGO dataset is documented in section 4 of topGO's the user manual:
https://bioconductor.org/packages/release/bioc/vignettes/topGO/inst/doc/topGO.pdf.
I need to use the `new`
command, and fill up the slots. The `annot` object must be a function that compiles "a list of
GO terms such that each element in the list is a character vector containing all the gene 
identifiers that are mapped to the respective GO term." There are several options, that you can
check using `help(annFUN.gene2GO)`, for example. The `annFUN.gene2GO` requires a `gene2GO`
argument, which is the list of gene-to-GO terms I made before. The `geneSelectionFun` object is a
function that selects the *interesting* (most significant) genes. It is required to
perform Fisher's exact test. The `nodeSize` is used to prune the GO hierarchy from the terms
which have less than `n` annotated genes.

I generate three datasets, to analyse each of the three ontologies.

```{r GOdata1, results = 'hold', message = FALSE}
selection <- function(allScores) {return(allScores < 0.05)}
GOdata.BP <- new('topGOdata',
   description = 'Differential expression among Brachionus plicatilis populations.',
   ontology = 'BP', 
   allGenes = PValues,
   annot = annFUN.gene2GO,
   gene2GO = allgenes2GO,
   geneSelectionFun = selection,
   nodeSize = 5)

GOdata.MF <- new('topGOdata',
   description = 'Differential expression among Brachionus plicatilis populations.',
   ontology = 'MF',
   allGenes = PValues,
   annot = annFUN.gene2GO,
   gene2GO = allgenes2GO,
   geneSelectionFun = selection,
   nodeSize = 5)

GOdata.CC <- new('topGOdata',
   description = 'Differential expression among Brachionus plicatilis populations.',
   ontology = 'CC',
   allGenes = PValues,
   annot = annFUN.gene2GO,
   gene2GO = allgenes2GO,
   geneSelectionFun = selection,
   nodeSize = 5)

DataSummary <- data.frame(ontology = c('BP', 'MF', 'CC'),
   Num_Genes = sapply(list(GOdata.BP, GOdata.MF, GOdata.CC), numGenes),
   Num_GO_terms = sapply(list(GOdata.BP, GOdata.MF, GOdata.CC), function(x) length(usedGO(x))))
kable(DataSummary, caption='Number of feasible genes or transcripts and number of GO terms used in each data set.')
rm(DataSummary)
```

## Running the tests

There are more than one way to test for enrichment. Something that took me a while to understand
is that not only there are different test statistics (Fisher's exact test, Kolmogorov-Smirnov,
and others) but also different *algorithms*: classic, elim, weight... The algorithms are ways
to deal with the dependence structure among GO terms due to topology. Some algorithms are
compatible with all statistics implemented in topGO. But the *weight* and the *parentchild*
algorithms are only applicable to Fisher's exact test. I am not interested in the classic
algorithm, which treats GO nodes as independent, and therefore produces an excess of significant
results. I will not use the Fisher's exact test, because it dependes on an arbitrary threshold
of significance on non-adjusted p-values.

```{r testing1, message = FALSE}
BP.elim     <- runTest(GOdata.BP, algorithm = "elim",     statistic = "ks")
BP.weight01 <- runTest(GOdata.BP, algorithm = "weight01", statistic = "ks")
BP.lea      <- runTest(GOdata.BP, algorithm = "lea",      statistic = "ks")
MF.elim     <- runTest(GOdata.MF, algorithm = "elim",     statistic = "ks")
MF.weight01 <- runTest(GOdata.MF, algorithm = "weight01", statistic = "ks")
MF.lea      <- runTest(GOdata.MF, algorithm = "lea",      statistic = "ks")
CC.elim     <- runTest(GOdata.CC, algorithm = "elim",     statistic = "ks")
CC.weight01 <- runTest(GOdata.CC, algorithm = "weight01", statistic = "ks")
CC.lea      <- runTest(GOdata.CC, algorithm = "lea",      statistic = "ks")

ResultsSummary <- data.frame(ontology = rep(c("BP", "MF", "CC"), each = 3),
   algorithm = rep(c("elim", "weight01", "lea"), 3),
   TermsTested = sapply(list(BP.elim, BP.weight01, BP.lea, MF.elim, MF.weight01, MF.lea, CC.elim, CC.weight01, CC.lea), function(X) length(score(X))),
   Significant = sapply(list(BP.elim, BP.weight01, BP.lea, MF.elim, MF.weight01, MF.lea, CC.elim, CC.weight01, CC.lea), function(X) sum(score(X) < 0.01)))

kable(ResultsSummary, caption="Number of non-trivial terms tested and those with a score (not corrected p-value) lower than 0.01.")
rm(ResultsSummary)
```

## Results

The `topGO` package does not pay much attention to what terms are significant because they are
*over*represented and which ones are *under*represented among the most differentially expressed
genes. I think it's worth separating them, to facilitate the biological interpretation. Note 
that not all terms listed in the tables below are significant. The scores for the three methods
(elim, weight01 and lea) are non-corrected p-values.

### Biological process

```{r BP1, message=FALSE}
orderedTerms <- names(sort(score(BP.weight01)))
significant.weight01 <- score(BP.weight01)[orderedTerms] <= 0.01
significant.lea      <- score(BP.lea)[orderedTerms] <= 0.01
significant.elim     <- score(BP.elim)[orderedTerms] <= 0.01
sigTerms <- orderedTerms[significant.weight01 & significant.lea & significant.elim]
# sigTerms gets updated, and is already used in the code, but I need a permanent copy for later.
BP.pvalue.sigTerms <- sigTerms

BP.all <- GenTable(GOdata.BP, elim=BP.elim, weight01=BP.weight01, lea=BP.lea,
                   orderBy="weight01", ranksOf="elim", topNodes=sum(significant.elim))

kable(
   BP.all[BP.all$Significant > BP.all$Expected,],
   caption = "Most over-represented terms of the Biological Process ontology.")

kable(
   BP.all[BP.all$Significant < BP.all$Expected,],
   caption = "Most under-represented terms of the Biological Process ontology.")

vennDiagram(vennCounts(cbind(weight01=significant.weight01,
                             lea=significant.lea,
                             elim=significant.elim)))

kable(data.frame(Term = Term(GOTERM[sigTerms]),
                 Definition = Definition(GOTERM[sigTerms]),
                 PValue=score(BP.weight01)[sigTerms]),
      caption = paste('Biological process terms significantly associated with',
                      VAR, 'according to all 3 algorithms', sep=' '))
```

I think the GO graph is useful to see the relationship among the significant terms. But
too large graphs are impossible to read. I don't know how to split the graph in meaningful
subgraphs.

```{r BP1dag, message=FALSE, fig.width=30, fig.height=15}
showSigOfNodes(GOdata.BP, score(BP.weight01),
               firstSigNodes = sum(significant.elim),
               wantedNodes = sigTerms)
```

This is just a example of the most significant GO term:

```{r BP1density, message=FALSE}
showGroupDensity(GOdata.BP, names(score(BP.weight01))[which.min(score(BP.weight01))])
```

### Molecular function

```{r MF1, message=FALSE}
orderedTerms <- names(sort(score(MF.weight01)))
significant.weight01 <- score(MF.weight01)[orderedTerms] <= 0.01
significant.lea      <- score(MF.lea)[orderedTerms] <= 0.01
significant.elim     <- score(MF.elim)[orderedTerms] <= 0.01
sigTerms <- orderedTerms[significant.weight01 & significant.lea & significant.elim]
# sigTerms gets updated, and is already used in the code, but I need a permanent copy for later.
MF.pvalue.sigTerms <- sigTerms

MF.all <- GenTable(GOdata.MF, elim=MF.elim, weight01=MF.weight01, lea=MF.lea,
                   orderBy="weight01", ranksOf="elim", topNodes=sum(significant.elim))

kable(
   MF.all[MF.all$Significant > MF.all$Expected,],
   caption = "Most over-represented terms of the Molecular Function ontology.")

kable(
   MF.all[MF.all$Significant < MF.all$Expected,],
   caption = "Most under-represented terms of the Molecular Function ontology.")

vennDiagram(vennCounts(cbind(weight01=significant.weight01,
                             lea=significant.lea,
                             elim=significant.elim)))

kable(data.frame(Term=Term(GOTERM[sigTerms]),
                 Definition=Definition(GOTERM[sigTerms]),
                 PValue=score(MF.weight01)[sigTerms]),
      caption = paste('Molecular function terms significantly associated with', VAR,
                      'according to all 3 algorithms', sep=' '))
```

```{r MF1dag, message=FALSE, fig.width=30, fig.height=15}
showSigOfNodes(GOdata.MF, score(MF.weight01),
               firstSigNodes = sum(significant.elim),
               wantedNodes = sigTerms)
```

```{r MF1density, message=FALSE}
showGroupDensity(GOdata.MF, names(score(MF.weight01))[which.min(score(MF.weight01))])
```

### Cellular component

```{r CC1, message=FALSE}
orderedTerms <- names(sort(score(CC.weight01)))
significant.weight01 <- score(CC.weight01)[orderedTerms] <= 0.01
significant.lea      <- score(CC.lea)[orderedTerms] <= 0.01
significant.elim     <- score(CC.elim)[orderedTerms] <= 0.01
sigTerms <- orderedTerms[significant.weight01 & significant.lea & significant.elim]
# sigTerms gets updated, and is already used in the code, but I need a permanent copy for later.
CC.pvalue.sigTerms <- sigTerms

CC.all <- GenTable(GOdata.CC, elim=CC.elim, weight01=CC.weight01, lea=CC.lea,
                   orderBy="weight01", ranksOf="elim", topNodes=sum(significant.elim))
kable(
   CC.all[CC.all$Significant > CC.all$Expected,],
   caption = "Most over-represented terms of the Cellular Component ontology.")

kable(
   CC.all[CC.all$Significant < CC.all$Expected,],
   caption = "Most under-represented terms of the Cellular Component ontology.")

vennDiagram(vennCounts(cbind(weight01=significant.weight01,
                             lea=significant.lea,
                             elim=significant.elim)))

kable(data.frame(Term=Term(GOTERM[sigTerms]),
                 Definition=Definition(GOTERM[sigTerms]),
                 PValue=score(CC.weight01)[sigTerms]),
      caption = paste('Cellular component terms significantly associated with', VAR,
                      'according to all 3 algorithms', sep=' '))
```

```{r CC1dag, message=FALSE, fig.width=30, fig.height=15}
showSigOfNodes(GOdata.CC, score(CC.weight01),
               firstSigNodes = sum(significant.elim),
               wantedNodes = sigTerms)
```

```{r CC1density, message=FALSE}
showGroupDensity(GOdata.CC, names(score(CC.weight01))[which.min(score(CC.weight01))])
```

# Using the portion of variance explained by `r VAR`
## Building the topGO object

I need to generate the topGO objects again, using the alternative gene ordering, based
on the proportion of expression-level variance explained by `r VAR`. I miss a way to
inform the `topGOdata` object that the score in this case is reversed, with respect to
p-values: the higher it is, the more differentially expressed the gene is. To make sure
that GO terms are tested in the same way than when using p-values, I will just reverse
the proportion of variance explained by `r VAR` to its complement. Taking this into
account, the subset of *interesting* genes (`selection()` function) must be defined as
the lowest 10% scores, which are the 10% genes with largest portion of variance explained
by `r VAR`.

```{r GOdata2, results = 'hold', message = FALSE}
selection <- function(allScores) {return(allScores <= quantile(allScores, probs = 0.10))}
GOdataVar.BP <- new('topGOdata',
   description = 'Differential expression among Brachionus plicatilis populations.',
   ontology = 'BP', 
   allGenes = 1.0 - Variance,
   annot = annFUN.gene2GO,
   gene2GO = allgenes2GO,
   geneSelectionFun = selection,
   nodeSize = 5)

GOdataVar.MF <- new('topGOdata',
   description = 'Differential expression among Brachionus plicatilis populations.',
   ontology = 'MF',
   allGenes = 1.0 - Variance,
   annot = annFUN.gene2GO,
   gene2GO = allgenes2GO,
   geneSelectionFun = selection,
   nodeSize = 5)

GOdataVar.CC <- new('topGOdata',
   description = 'Differential expression among Brachionus plicatilis populations.',
   ontology = 'CC',
   allGenes = 1.0 - Variance,
   annot = annFUN.gene2GO,
   gene2GO = allgenes2GO,
   geneSelectionFun = selection,
   nodeSize = 5)

library(knitr)
DataSummary <- data.frame(ontology = c('BP', 'MF', 'CC'),
   Num_Genes = sapply(list(GOdataVar.BP, GOdataVar.MF, GOdataVar.CC), numGenes),
   Num_GO_terms = sapply(list(GOdataVar.BP, GOdataVar.MF, GOdataVar.CC), function(x) length(usedGO(x))))
kable(DataSummary, caption='Number of feasible genes or transcripts and number of GO terms used in each data set.')
rm(DataSummary)
```

## Running the tests

```{r testing2, message = FALSE}
BPvar.elim     <- runTest(GOdataVar.BP, algorithm = "elim",     statistic = "ks")
BPvar.weight01 <- runTest(GOdataVar.BP, algorithm = "weight01", statistic = "ks")
BPvar.lea      <- runTest(GOdataVar.BP, algorithm = "lea",      statistic = "ks")
MFvar.elim     <- runTest(GOdataVar.MF, algorithm = "elim",     statistic = "ks")
MFvar.weight01 <- runTest(GOdataVar.MF, algorithm = "weight01", statistic = "ks")
MFvar.lea      <- runTest(GOdataVar.MF, algorithm = "lea",      statistic = "ks")
CCvar.elim     <- runTest(GOdataVar.CC, algorithm = "elim",     statistic = "ks")
CCvar.weight01 <- runTest(GOdataVar.CC, algorithm = "weight01", statistic = "ks")
CCvar.lea      <- runTest(GOdataVar.CC, algorithm = "lea",      statistic = "ks")

ResultsSummary <- data.frame(ontology = rep(c("BP", "MF", "CC"), each = 3),
   algorithm = rep(c("elim", "weight01", "lea"), 3),
   TermsTested = sapply(list(BPvar.elim, BPvar.weight01, BPvar.lea, MFvar.elim, MFvar.weight01, MFvar.lea, CCvar.elim, CCvar.weight01, CCvar.lea), function(X) length(score(X))),
   Significant = sapply(list(BPvar.elim, BPvar.weight01, BPvar.lea, MFvar.elim, MFvar.weight01, MFvar.lea, CCvar.elim, CCvar.weight01, CCvar.lea), function(X) sum(score(X) < 0.01)))

kable(ResultsSummary, caption="Number of non-trivial terms tested and those with a score (not corrected p-value) lower than 0.01.")
rm(ResultsSummary)
```

## Results
### Biological process

```{r BP2, message=FALSE}
orderedTerms <- names(sort(score(BPvar.weight01)))
significant.weight01 <- score(BPvar.weight01)[orderedTerms] <= 0.01
significant.lea      <- score(BPvar.lea)[orderedTerms] <= 0.01
significant.elim     <- score(BPvar.elim)[orderedTerms] <= 0.01
sigTerms <- orderedTerms[significant.weight01 & significant.lea & significant.elim]
# sigTerms gets updated, and is already used in the code, but I need a permanent copy for later.
BP.variance.sigTerms <- sigTerms

BPvar.all <- GenTable(GOdataVar.BP, elim=BPvar.elim, weight01=BPvar.weight01, lea=BPvar.lea,
                      orderBy="weight01", ranksOf="elim", topNodes=sum(significant.elim))

kable(
   BPvar.all[BPvar.all$Significant > BPvar.all$Expected,],
   caption = "Most over-represented terms of the Biological Process ontology.")

kable(
   BPvar.all[BPvar.all$Significant < BPvar.all$Expected,],
   caption = "Most under-represented terms of the Biological Process ontology.")

vennDiagram(vennCounts(cbind(weight01=significant.weight01,
                             lea=significant.lea,
                             elim=significant.elim)))

kable(data.frame(Term=Term(GOTERM[sigTerms]),
                 Definition=Definition(GOTERM[sigTerms]),
                 PValue=score(BPvar.weight01)[sigTerms]),
      caption = paste('Biological process terms significantly associated with', VAR,
                      'according to all 3 algorithms', sep=' '))
```

```{r BP2dag, message=FALSE, fig.width=30, fig.height=15}
showSigOfNodes(GOdataVar.BP, score(BPvar.weight01),
               firstSigNodes = sum(significant.elim),
               wantedNodes = sigTerms)
```

Below I plot variance portion, but for the termo found most significant when using
p-values, for comparison.

```{r BP2density, message=FALSE}
showGroupDensity(GOdataVar.BP, names(score(BP.weight01))[which.min(score(BP.weight01))])
```

### Molecular function

```{r MF2, message=FALSE}
orderedTerms <- names(sort(score(MFvar.weight01)))
significant.weight01 <- score(MFvar.weight01)[orderedTerms] <= 0.01
significant.lea      <- score(MFvar.lea)[orderedTerms] <= 0.01
significant.elim     <- score(MFvar.elim)[orderedTerms] <= 0.01
sigTerms <- orderedTerms[significant.weight01 & significant.lea & significant.elim]
# sigTerms gets updated, and is already used in the code, but I need a permanent copy for later.
MF.variance.sigTerms <- sigTerms

MFvar.all <- GenTable(GOdataVar.MF, elim=MFvar.elim, weight01=MFvar.weight01, lea=MFvar.lea,
                      orderBy="weight01", ranksOf="elim", topNodes=sum(significant.elim))
kable(
   MFvar.all[MFvar.all$Significant > MFvar.all$Expected,],
   caption = "Most over-represented terms of the Molecular Function ontology.")

kable(
   MFvar.all[MFvar.all$Significant < MFvar.all$Expected,],
   caption = "Most under-represented terms of the Molecular Function ontology.")

vennDiagram(vennCounts(cbind(weight01=significant.weight01,
                             lea=significant.lea,
                             elim=significant.elim)))

kable(data.frame(Term=Term(GOTERM[sigTerms]),
                 Definition=Definition(GOTERM[sigTerms]),
                 PValue=score(MFvar.weight01)[sigTerms]),
      caption = paste('Molecular functions terms significantly associated with', VAR,
                      'according to all 3 algorithms', sep=' '))
```

```{r MF2dag, message=FALSE, fig.width=30, fig.height=15}
showSigOfNodes(GOdataVar.MF, score(MFvar.weight01),
               firstSigNodes = sum(significant.elim),
               wantedNodes = sigTerms)
```

I plot variance portion, but for the term found most significant when using p-values, for comparison.

```{r MF2density, message=FALSE}
showGroupDensity(GOdataVar.MF, names(score(MF.weight01))[which.min(score(MF.weight01))])
```

### Cellular component

```{r CC2, message=FALSE}
orderedTerms <- names(sort(score(CCvar.weight01)))
significant.weight01 <- score(CCvar.weight01)[orderedTerms] <= 0.01
significant.lea      <- score(CCvar.lea)[orderedTerms] <= 0.01
significant.elim     <- score(CCvar.elim)[orderedTerms] <= 0.01
sigTerms <- orderedTerms[significant.weight01 & significant.lea & significant.elim]
if (length(sigTerms) == 0) {
   sigTerms <- orderedTerms[significant.lea & significant.elim]
}
# sigTerms gets updated, and is already used in the code, but I need a permanent copy for later.
CC.variance.sigTerms <- sigTerms

CCvar.all <- GenTable(GOdataVar.CC, elim=CCvar.elim, weight01=CCvar.weight01, lea=CCvar.lea,
                      orderBy="weight01", ranksOf="elim", topNodes=max(sum(significant.elim), 10))

kable(
   CCvar.all[CCvar.all$Significant > CCvar.all$Expected,],
   caption = "Most over-represented terms of the Cellular Component ontology.")

kable(
   CCvar.all[CCvar.all$Significant < CCvar.all$Expected,],
   caption = "Most under-represented terms of the Cellular Component ontology.")

vennDiagram(vennCounts(cbind(weight01=significant.weight01,
                             lea=significant.lea,
                             elim=significant.elim)))

kable(data.frame(Term=Term(GOTERM[sigTerms]),
                 Definition=Definition(GOTERM[sigTerms]),
                 PValue=score(CCvar.weight01)[sigTerms]),
      caption = paste('Cellular component terms significantly associated with', VAR,
                      'according to all 3 algorithms', sep=' '))
```

```{r CC2dag, message=FALSE, fig.width=30, fig.height=15}
showSigOfNodes(GOdataVar.CC, score(CCvar.weight01),
               firstSigNodes = sum(significant.weight01),
               wantedNodes = sigTerms)
```

For comparison, I plot the distribution of variance portion for the CC term found most significant when using p-values.

```{r CC2density, message=FALSE}
showGroupDensity(GOdataVar.CC, names(score(CC.weight01))[which.min(score(CC.weight01))])
```

# Comparison between the two ordering of genes.
## Biological process

```{r BPcomparison, message=FALSE}
allTerms <- usedGO(GOdata.BP)
vennDiagram(vennCounts(cbind(PValue   = allTerms %in% BP.pvalue.sigTerms,
                             Variance = allTerms %in% BP.variance.sigTerms)))

ggplot(data = data.frame(PValue   = rank(score(BP.weight01))[allTerms],
                         Variance = rank(score(BPvar.weight01))[allTerms]),
       mapping = aes(x = PValue, y = Variance)) +
   geom_point() + geom_smooth() + xlab('Using gene p-values') +
   ylab('Using portion of explained variance') +
   ggtitle('Ordering of BP terms by significance')
```

## Molecular function

```{r MFcomparison, message=FALSE}
allTerms <- usedGO(GOdata.MF)
vennDiagram(vennCounts(cbind(PValue   = allTerms %in% MF.pvalue.sigTerms,
                             Variance = allTerms %in% MF.variance.sigTerms)))

ggplot(data = data.frame(PValue   = rank(score(MF.weight01))[allTerms],
                         Variance = rank(score(MFvar.weight01))[allTerms]),
       mapping = aes(x = PValue, y = Variance)) +
   geom_point() + geom_smooth() + xlab('Using gene p-values') +
   ylab('Using portion of explained variance') +
   ggtitle('Ordering of MF terms by significance')
```

## Cellular component

```{r CCcomparison, message=FALSE}
allTerms <- usedGO(GOdata.CC)
vennDiagram(vennCounts(cbind(PValue   = allTerms %in% CC.pvalue.sigTerms,
                             Variance = allTerms %in% CC.variance.sigTerms)))

ggplot(data = data.frame(PValue   = rank(score(CC.weight01))[allTerms],
                         Variance = rank(score(CCvar.weight01))[allTerms]),
       mapping = aes(x = PValue, y = Variance)) +
   geom_point() + geom_smooth() + xlab('Using gene p-values') +
   ylab('Using portion of explained variance') +
   ggtitle('Ordering of CC terms by significance')
```

# Session info
I save the main variables to be able to load them in a new R session later.
```{r last}
save(allgenes2GO,
     GOdata.BP, BP.elim, BP.weight01, BP.lea, BP.pvalue.sigTerms,
     GOdata.MF, MF.elim, MF.weight01, MF.lea, MF.pvalue.sigTerms,
     GOdata.CC, CC.elim, CC.weight01, CC.lea, CC.pvalue.sigTerms,
     GOdataVar.BP, BPvar.elim, BPvar.weight01, BPvar.lea, BP.variance.sigTerms,
     GOdataVar.MF, MFvar.elim, MFvar.weight01, MFvar.lea, MF.variance.sigTerms,
     GOdataVar.CC, CCvar.elim, CCvar.weight01, CCvar.lea, CC.variance.sigTerms,
     file = paste('Enrichment', TAG, VAR, 'RData', sep='.'))
sessionInfo()
```