---
title: "Effect of hatching condition on gene expression"
output: html_document
author: "J. Ignacio Lucas Lledó"
---

```{r eval=TRUE, echo=FALSE, include=FALSE}
library(edgeR)
library(ggplot2)
args <- commandArgs(trailingOnly = TRUE)
if (length(args) == 0) {
   stop("At least one argument must be supplied: input file.", call.=FALSE)
} else if (length(args) == 1) {
   args[2] <- "."
}

input_file_name <- args[1]
output_dir <- args[2]

population <- factor(c("1","1","2","2","3","3","4","4","5","5","6","6"))
NestedPop  <- factor(c("1","1","2","2","1","1","3","3","2","2","3","3"))
diapause   <- factor(rep(c("hatch","diapause"), 6))
diapause   <- relevel(diapause, "hatch")
regime     <- factor(c("random","random","random","random","regular","regular",
                 "random","random","regular","regular","regular","regular"))
regime     <- relevel(regime, "regular")
combined   <- factor(c("RandHatch","RandDiap","RandHatch","RandDiap","RegulHatch","RegulDiap",
                      "RandHatch","RandDiap","RegulHatch","RegulDiap","RegulHatch","RegulDiap"))
combined   <- relevel(combined, "RegulHatch")

if (!exists('y')) {
   if (!exists('counts')) {
      counts <- read.table(input_file_name, row.names=1, header=TRUE)
   }
   y <- DGEList(counts=counts, group=regime)
   y <- calcNormFactors(y, method="TMM")
   threshold <- 5.0 / (min(y$samples$lib.size) / 1000000)
   keep <- rowSums(cpm(y) > threshold) >= 4
   y <- y[keep, ,keep.lib.sizes=FALSE]
}

NumOfTags <- dim(y)[1]

if (!exists('fit1')) {
   design1 <- model.matrix(~population + diapause)
   if (!exists('y1')) {
      y1 <- estimateDisp(y, design1)
   }
   fit1 <- glmQLFit(y1, design1)
}

if (!exists('fit2')) {
   design2 <- model.matrix(~regime + diapause)
   if (!exists('y2')) {
      y2 <- estimateDisp(y, design2)
   }
   fit2 <- glmQLFit(y2, design2)   
}

if (!exists('fit3')) {
   design3 <- model.matrix(~0 + combined)
   if (!exists('y3')) {
      y3 <- estimateDisp(y, design3)
   }
   fit3 <- glmQLFit(y3, design3)
}

if (!exists('fit4')) {
   design4 <- model.matrix(~regime + diapause + regime:NestedPop)
   if (!exists('y4')) {
      y4 <- estimateDisp(y, design4)
   }
   fit4 <- glmQLFit(y4, design4)
}

if (!exists('fit5')) {
   design5 <- model.matrix(~regime + diapause + regime:diapause)
   if (!exists('y5')) {
      y5 <- estimateDisp(y, design5)
   }
   fit5 <- glmQLFit(y5, design5)
}

if (!exists('fit6')) {
   design6 <- model.matrix(~regime + diapause + regime:diapause + regime:NestedPop)
   if (!exists('y6')) {
      y6 <- estimateDisp(y, design6)
   }
   fit6 <- glmQLFit(y6, design6)
}

if (!exists('I1vsI2')) {
   I1vsI2 <- read.table("../2017-10-27/list_genes/I1vsI2_all.txt", 
      colClasses=c("NULL", rep("character", 6), rep("numeric", 6),"factor"),
      col.names=c("ignored","geneid","gene","locus","sample1","sample2","testStat",
      "FPKMx","FPKMy","log2FC","statistic","p","q","significant"))
}

if (!exists('P1vsP2')) {
   P1vsP2 <- read.table("../2017-10-27/list_genes/P1vsP2_all.txt",
      colClasses=c("NULL", rep("character", 6), rep("numeric", 6),"factor"),
      col.names=c("ignored","geneid","gene","locus","sample1","sample2","testStat",
      "FPKMx","FPKMy","log2FC","statistic","p","q","significant"))
}

if (!exists('I1vsP1')) {
   I1vsP1 <- read.table("../2017-10-27/list_genes/I1vsP1_all.txt",
      colClasses=c("NULL", rep("character", 6), rep("numeric", 6),"factor"),
      col.names=c("ignored","geneid","gene","locus","sample1","sample2","testStat",
      "FPKMx","FPKMy","log2FC","statistic","p","q","significant"))
}

if (!exists('I2vsP2')) {
   I2vsP2 <- read.table("../2017-10-27/list_genes/I2vsP2_all.txt",
      colClasses=c("NULL", rep("character", 6), rep("numeric", 6),"factor"),
      col.names=c("ignored","geneid","gene","locus","sample1","sample2","testStat",
      "FPKMx","FPKMy","log2FC","statistic","p","q","significant"))
}

if (!exists('Cuffdiff_hatching')) {
   Cuffdiff_hatching <- I1vsI2$geneid[I1vsI2$geneid %in% P1vsP2$geneid]
}

if (!exists('Cuffdiff_regime')) {
   Cuffdiff_regime   <- I1vsP1$geneid[I1vsP1$geneid %in% I2vsP2$geneid]
}
```
# Model 1

Model 1 is expected to be the most sensitive to detect genes the expression of which is
affected by enforcing diapause, after accounting for the variation among populations (which
includes variation due to selective regime).

```{r}
qlf1 <- glmQLFTest(fit1)
topTags(qlf1)
z1 <- topTags(qlf1, n=NumOfTags)$table
z1$logPValue <- -log10(z1$PValue)
ggplot(data=z1, mapping=aes(x=logFC, y=logPValue, color=FDR)) +
   geom_point() + theme_grey() + ylab(expression(paste("-log"[10], plain("(p value)")))) +
   ggtitle('Effect of enforcing diapause, model 1')
```

I would say the two first genes (XLOC_052427 and XLOC_045087) are significantly downregulated
when enforcing diapause, with respect to the immediate hatching condition.

# Model 2

Model 2 is very simple, and may include false positives, because it does not account for any
interaction, nor for among-population variance.

```{r}
qlf2 <- glmQLFTest(fit2, coef=3)
topTags(qlf2)
z1 <- topTags(qlf2, n=NumOfTags)$table
z1$logPValue <- -log10(z1$PValue)
ggplot(data=z1, mapping=aes(x=logFC, y=logPValue, color=FDR)) +
   geom_point() + theme_grey() + ylab(expression(paste("-log"[10], plain("(p value)")))) +
   ggtitle('Effect of enforcing diapause, model 2')
```

According to model 2, there are `r dim(topTags(qlf2, n=NumOfTags, p.value=0.10))[1]` tags
significantly up- or downregulated by hatching condition, at FDR=0.1.

# Model 3

In model 3, the effect of diapause may be understood in two different ways. First, we may want
to know what genes have an average expression level different between hatching conditions. This
is different from testing the effect of hatching condition twice: first among samples in a regular
environment and then among those in the stochastic environment. Then, I could select the genes
with the same significant tendency in both selective regimes. I will start this analysis to the
first option, comparing only average gene expression level.

```{r}
qlf3 <- glmQLFTest(fit3, contrast=c(-0.5, 0.5, -0.5, 0.5))
topTags(qlf3)
z1 <- topTags(qlf3, n=NumOfTags)$table
z1$logPValue <- -log10(z1$PValue)
ggplot(data=z1, mapping=aes(x=logFC, y=logPValue, color=FDR)) +
   geom_point() + theme_grey() + ylab(expression(paste("-log"[10], plain("(p value)")))) +
   ggtitle('Effect of enforcing diapause, model 3')
```

According to model 3, there are `r dim(topTags(qlf3, n=NumOfTags, p.value=0.10))[1]` tags
significantly up- or downregulated by hatching condition at FDR = 0.1.

For the sake of comparing model 3 to the results from Cuffdiff, I will make each one of the
four possible comparisons. The four coefficients of model correspond, in this order, to:
regular regime and immediate hatching, random regime and diapause, random regime and hatching,
and regular regime and diapause. In Eva's terms, I believe they correspond to P1, I2, I1, and
P2.

```{r fig.width=6, fig.height=6}
#                                     P1 I2 I1 P2
qlf3_1 <- glmQLFTest(fit3, contrast=c( 0,-1, 1, 0)) # I1vsI2
qlf3_2 <- glmQLFTest(fit3, contrast=c( 1, 0, 0,-1)) # P1vsP2
qlf3_3 <- glmQLFTest(fit3, contrast=c(-1, 0, 1, 0)) # I1vsP1
qlf3_4 <- glmQLFTest(fit3, contrast=c( 0, 1, 0,-1)) # I2vsP2

z1 <- topTags(qlf3_1, n=NumOfTags)$table
z2 <- topTags(qlf3_2, n=NumOfTags)$table
z3 <- topTags(qlf3_3, n=NumOfTags)$table
z4 <- topTags(qlf3_4, n=NumOfTags)$table

z1$inCuffdiff <- row.names(z1) %in% I1vsI2$geneid
z2$inCuffdiff <- row.names(z2) %in% P1vsP2$geneid
z3$inCuffdiff <- row.names(z3) %in% I1vsP1$geneid
z4$inCuffdiff <- row.names(z4) %in% I2vsP2$geneid

layout(mat=matrix(c(1,2,3,4), ncol=2, nrow=2, byrow=TRUE))
qqplot(z1$PValue[z1$inCuffdiff], z1$PValue[!z1$inCuffdiff], xlab='Cuffdiff genes p-values', ylab='Other genes p-values', main='Hatching vs Diapause, when Random')
qqplot(z2$PValue[z2$inCuffdiff], z2$PValue[!z2$inCuffdiff], xlab='Cuffdiff genes p-values', ylab='Other genes p-values', main='Hatching vs Diapause, when Regular')
qqplot(z3$PValue[z3$inCuffdiff], z3$PValue[!z3$inCuffdiff], xlab='Cuffdiff genes p-values', ylab='Other genes p-values', main='Random vs Regular, when hatching')
qqplot(z4$PValue[z4$inCuffdiff], z4$PValue[!z4$inCuffdiff], xlab='Cuffdiff genes p-values', ylab='Other genes p-values', main='Random vs Regular, when diapausing')
```

The graphs above compare the distribution of p-values of genes called differentially regulated by
Cuffdiff with the distribution of p-values among the rest of genes. It can be seen that in all
comparisons, the genes detected as differentially expressed by Cuffdiff are not assigned lower
p-values by EdgeR in the same comparisons. Rather, they look random samples of genes with respect
to their p-values. 

# Model 4

```{r}
qlf4 <- glmQLFTest(fit4, coef=3)
topTags(qlf4)
z1 <- topTags(qlf4, n=NumOfTags)$table
z1$logPValue <- -log10(z1$PValue)
ggplot(data=z1, mapping=aes(x=logFC, y=logPValue, color=FDR)) +
   geom_point() + theme_grey() + ylab(expression(paste("-log"[10], plain("(p value)")))) +
   ggtitle('Effect of enforcing diapause, model 4')
```

According to model 4, there are `r dim(topTags(qlf4, n=NumOfTags, p.value=0.10))[1]` tags
significantly up- or downregulated by hatching condition at FDR = 0.1.

# Model 5

```{r}
qlf5 <- glmQLFTest(fit5, coef=3)
topTags(qlf5)
z1 <- topTags(qlf5, n=NumOfTags)$table
z1$logPValue <- -log10(z1$PValue)
ggplot(data=z1, mapping=aes(x=logFC, y=logPValue, color=FDR)) +
   geom_point() + theme_grey() + ylab(expression(paste("-log"[10], plain("(p value)")))) +
   ggtitle('Effect of enforcing diapause, model 5')
```

According to model 5, there are `r dim(topTags(qlf5, n=NumOfTags, p.value=0.10))[1]` tags
significantly up- or downregulated by hatching condition at FDR = 0.1.

# Model 6

```{r}
qlf6 <- glmQLFTest(fit6, coef=3)
topTags(qlf6)
z1 <- topTags(qlf6, n=NumOfTags)$table
z1$logPValue <- -log10(z1$PValue)
ggplot(data=z1, mapping=aes(x=logFC, y=logPValue, color=FDR)) +
   geom_point() + theme_grey() + ylab(expression(paste("-log"[10], plain("(p value)")))) +
   ggtitle('Effect of enforcing diapause, model 6')
```

According to model 6, there are `r dim(topTags(qlf6, n=NumOfTags, p.value=0.10))[1]` tags
significantly up- or downregulated by hatching condition at FDR = 0.10. And this is how
the results from models 1, 2, 3 and 4 match:

```{r}
significant <- cbind(model1 = topTags(qlf1, n=NumOfTags, sort.by='none')$table$FDR <= 0.4, 
                     model2 = topTags(qlf2, n=NumOfTags, sort.by='none')$table$FDR <= 0.4,
                     model3 = topTags(qlf3, n=NumOfTags, sort.by='none')$table$FDR <= 0.4,
                     model4 = topTags(qlf4, n=NumOfTags, sort.by='none')$table$FDR <= 0.4)
vennDiagram(vennCounts(significant), main='FDR=0.4')
```

The genes detected at FDR < 0.4 by models 1, 2 are
`r row.names(topTags(qlf1,p.value=0.4))[row.names(topTags(qlf1,p.value=0.4)) %in% row.names(topTags(qlf2,p.value=0.4))]`
. No need for a functional enrichment analysis. 
