---
title: "Testing the interactions terms in the models of gene expression"
output: html_document
author: "J. Ignacio Lucas Lledó"
---

```{r eval=TRUE, echo=FALSE, include=FALSE}
library(edgeR)
library(ggplot2)
args <- commandArgs(trailingOnly = TRUE)
if (length(args) == 0) {
   stop("At least one argument must be supplied: input file.", call.=FALSE)
} else if (length(args) == 1) {
   args[2] <- "."
}

input_file_name <- args[1]
output_dir <- args[2]

population <- factor(c("1","1","2","2","3","3","4","4","5","5","6","6"))
NestedPop  <- factor(c("1","1","2","2","1","1","3","3","2","2","3","3"))
diapause   <- factor(rep(c("hatch","diapause"), 6))
diapause   <- relevel(diapause, "hatch")
regime     <- factor(c("random","random","random","random","regular","regular",
                 "random","random","regular","regular","regular","regular"))
regime     <- relevel(regime, "regular")
combined   <- factor(c("RandHatch","RandDiap","RandHatch","RandDiap","RegulHatch","RegulDiap",
                      "RandHatch","RandDiap","RegulHatch","RegulDiap","RegulHatch","RegulDiap"))
combined   <- relevel(combined, "RegulHatch")

if (!exists('y')) {
   if (!exists('counts')) {
      counts <- read.table(input_file_name, row.names=1, header=TRUE)
   }
   y <- DGEList(counts=counts, group=regime)
   y <- calcNormFactors(y, method="TMM")
   threshold <- 5.0 / (min(y$samples$lib.size) / 1000000)
   keep <- rowSums(cpm(y) > threshold) >= 4
   y <- y[keep, ,keep.lib.sizes=FALSE]
}

NumOfTags <- dim(y)[1]

if (!exists('fit1')) {
   design1 <- model.matrix(~population + diapause)
   if (!exists('y1')) {
      y1 <- estimateDisp(y, design1)
   }
   fit1 <- glmQLFit(y1, design1)
}

if (!exists('fit2')) {
   design2 <- model.matrix(~regime + diapause)
   if (!exists('y2')) {
      y2 <- estimateDisp(y, design2)
   }
   fit2 <- glmQLFit(y2, design2)   
}

if (!exists('fit3')) {
   design3 <- model.matrix(~0 + combined)
   if (!exists('y3')) {
      y3 <- estimateDisp(y, design3)
   }
   fit3 <- glmQLFit(y3, design3)
}

if (!exists('fit4')) {
   design4 <- model.matrix(~regime + diapause + regime:NestedPop)
   if (!exists('y4')) {
      y4 <- estimateDisp(y, design4)
   }
   fit4 <- glmQLFit(y4, design4)
}

if (!exists('fit5')) {
   design5 <- model.matrix(~regime + diapause + regime:diapause)
   if (!exists('y5')) {
      y5 <- estimateDisp(y, design5)
   }
   fit5 <- glmQLFit(y5, design5)
}

if (!exists('fit6')) {
   design6 <- model.matrix(~regime + diapause + regime:diapause + regime:NestedPop)
   if (!exists('y6')) {
      y6 <- estimateDisp(y, design6)
   }
   fit6 <- glmQLFit(y6, design6)
}

if (!exists('I1vsI2')) {
   I1vsI2 <- read.table("../2017-10-27/list_genes/I1vsI2_all.txt", 
      colClasses=c("NULL", rep("character", 6), rep("numeric", 6),"factor"),
      col.names=c("ignored","geneid","gene","locus","sample1","sample2","testStat",
      "FPKMx","FPKMy","log2FC","statistic","p","q","significant"))
}

if (!exists('P1vsP2')) {
   P1vsP2 <- read.table("../2017-10-27/list_genes/P1vsP2_all.txt",
      colClasses=c("NULL", rep("character", 6), rep("numeric", 6),"factor"),
      col.names=c("ignored","geneid","gene","locus","sample1","sample2","testStat",
      "FPKMx","FPKMy","log2FC","statistic","p","q","significant"))
}

if (!exists('I1vsP1')) {
   I1vsP1 <- read.table("../2017-10-27/list_genes/I1vsP1_all.txt",
      colClasses=c("NULL", rep("character", 6), rep("numeric", 6),"factor"),
      col.names=c("ignored","geneid","gene","locus","sample1","sample2","testStat",
      "FPKMx","FPKMy","log2FC","statistic","p","q","significant"))
}

if (!exists('I2vsP2')) {
   I2vsP2 <- read.table("../2017-10-27/list_genes/I2vsP2_all.txt",
      colClasses=c("NULL", rep("character", 6), rep("numeric", 6),"factor"),
      col.names=c("ignored","geneid","gene","locus","sample1","sample2","testStat",
      "FPKMx","FPKMy","log2FC","statistic","p","q","significant"))
}

if (!exists('Cuffdiff_hatching')) {
   Cuffdiff_hatching <- I1vsI2$geneid[I1vsI2$geneid %in% P1vsP2$geneid]
}

if (!exists('Cuffdiff_regime')) {
   Cuffdiff_regime   <- I1vsP1$geneid[I1vsP1$geneid %in% I2vsP2$geneid]
}
```
# Interaction between selective regime and hatching condition
## Model 6

```{r}
qlf6 <- glmQLFTest(fit6, coef=4)
topTags(qlf6)
z1 <- topTags(qlf6, n=NumOfTags)$table
z1$logPValue <- -log10(z1$PValue)
ggplot(data=z1, mapping=aes(x=logFC, y=logPValue, color=FDR)) + 
   geom_point() + theme_grey() + ylab(expression(paste("-log"[10], plain("(p value)")))) +
   ggtitle('Interaction regime-hatching')
```

According to model 6, there are `r dim(topTags(qlf6, n=NumOfTags, p.value=0.25))[1]` tags
with a significant interaction term between selective regime and hatching condition at
FDR = 0.25.

## Model 5

```{r}
qlf5 <- glmQLFTest(fit5, coef=4)
topTags(qlf5)
z1 <- topTags(qlf5, n=NumOfTags)$table
z1$logPValue <- -log10(z1$PValue)
ggplot(data=z1, mapping=aes(x=logFC, y=logPValue, color=FDR)) +
   geom_point() + theme_grey() + ylab(expression(paste("-log"[10], plain("p value)")))) + 
   ggtitle('Interaction regime-hatching in model 5')
```

According to model 5, there are `r dim(topTags(qlf5, n=NumOfTags, p.value=0.25))[1]` tags
with a significant interaction term between selective regime and hatching condition at 
FDR = 0.25. And this is how models 5 and 6 agree on what genes have such significant interaction
term:

```{r}
significant <- cbind(model5 = topTags(qlf5, n=NumOfTags, sort.by='none')$table$FDR <= 0.25,
                     model6 = topTags(qlf6, n=NumOfTags, sort.by='none')$table$FDR <= 0.25)
vennDiagram(vennCounts(significant))
```

# Interaction between selective regime and population

When testing for significant coefficients showing the effect of the nested population, there
are not one but 4 possible coefficients, each with a different fold change. For the volcano
plot, I want to use for each gene the fold change that is most extreme (either positive or
negative).

## Model 6

```{r}
qlf6  <- glmQLFTest(fit6, coef=5:8)
topTags(qlf6)
z1 <- topTags(qlf6, n=NumOfTags)$table
z1$logPValue <- -log10(z1$PValue)
z1$logFC <- z1[cbind(1:NumOfTags, apply(abs(z1[,1:4]), 1, which.max))]
ggplot(data=z1, mapping=aes(x=logFC, y=logPValue, color=FDR)) +
   geom_point() + theme_grey() + ylab(expression(paste("-log"[10], plain("p value)")))) +
   ggtitle('Interaction regime-population in complete model')
```

According to model 6, there are `r dim(topTags(qlf6, n=NumOfTags, p.value=0.01))[1]` tags
with a significant interaction term between selective regime and population, at FDR = 0.01.

## Model 4

```{r}
qlf4 <- glmQLFTest(fit4, coef=4:7)
topTags(qlf4)
z1 <- topTags(qlf4, n=NumOfTags)$table
z1$logPValue <- -log10(z1$PValue)
z1$logFC <- z1[cbind(1:NumOfTags, apply(abs(z1[,1:4]), 1, which.max))]
ggplot(data=z1, mapping=aes(x=logFC, y=logPValue, color=FDR)) +
   geom_point() + theme_grey() + ylab(expression(paste("-log"[10], plain("p value)")))) + 
   ggtitle('Interaction regime-population in model 4')
```

According to model 4, there are `r dim(topTags(qlf4, n=NumOfTags, p.value=0.01))[1]` tags
with a significant interaction term between selective regimew and population, at FDR = 0.01.
And this is how the genes identified as having such significant coefficients match between
results of models 4 and 6:

```{r}
significant <- cbind(model4 = topTags(qlf4, n=NumOfTags, sort.by='none')$table$FDR <= 0.01,
                     model6 = topTags(qlf6, n=NumOfTags, sort.by='none')$table$FDR <= 0.01)
vennDiagram(vennCounts(significant))
```
