---
title: "Differential gene expression analysis with EdgeR"
output: html_document
author: "J. Ignacio Lucas Lledó"
---

```{r eval=TRUE, echo=FALSE, include=FALSE}
library(edgeR)
library(ggplot2)
```
# Reading the data

The data for this analysis was created in folder 2019-03-29. 

```{r eval=TRUE, echo=TRUE, include=TRUE}
args <- commandArgs(trailingOnly = TRUE)
if (length(args) == 0) {
   stop("At least one argument must be supplied: input file.", call.=FALSE)
} else if (length(args) == 1) {
   args[2] <- "."
}

input_file_name <- args[1]
output_dir <- args[2]
counts <- read.table(input_file_name, row.names=1, header=TRUE)
```

Three factors: population, diapause or hatching condition, and selective regime.
I prefer the immediate hatching (instead of forced diapause) and the regular or
periodic selective regime as the reference levels of the corresponding factors.

```{r}
population  <- factor(c("1","1","2","2","1","1","3","3","2","2","3","3"))
diapause <- factor(rep(c("hatch","diapause"), 6))
diapause <- relevel(diapause, "hatch")
regime <- factor(c("random","random","random","random","regular","regular",
                 "random","random","regular","regular","regular","regular"))
regime <- relevel(regime, "regular")

y <- DGEList(counts=counts, group=regime)
y <- calcNormFactors(y, method="TMM")

barplot(y$samples$lib.size*1e-6, names=1:12, ylab="Library size (millions)")
```
It looks like the populations undergoing diapause systematically produced smaller
libraries. It may be the effect of a smaller amount of eggs available. Population 5
is the exception.


#  Filtering

The recommended filtering consists on removing genes that have such a low expression
level that do not have at least 5-10 counts on as many samples as groups there are. The
rationale is that those genes cannot have enough reads in at least one sample of each
group. However, instead of filtering on the raw counts, the edgeR manual suggests to
translate whatever minimum number of reads per library in terms of counts per million.
The reason is that the filtering should take into account differences in library sizes
between samples. It probably makes sense, because otherwise, I would be applying an
effectively higher (harsher) minimum threshold to the gene expression level in samples
with smaller libraries. That would probably bias results, since we would be enriching
the dataset in genes more highly expressed in some samples than in others.

Say that library sizes range between 250000 and 2000000. Then, applying a minimum expression
level of 20 counts per million is the only way to make sure that even the small libraries
are required to contain at least 5 counts, while at the same time setting the cutoff at the
same biologically relevant level. In any case, as long as the minimum expression
level is not enforced in all samples, the dataset retained will contain genes that do not
reach the minimum in some samples. Since counts per million are (expectedly and actually)
similarly distributed among samples, it should not be the case that any sample is enriched
in low-expression genes.

```{r}
threshold <- 5.0 / (min(y$samples$lib.size) / 1000000)
keep <- rowSums(cpm(y) > threshold) >= 4
y <- y[keep, ,keep.lib.sizes=FALSE]

plotMDS(y$counts)
```

The magnitude of variation along the axes seems very high in comparison with the examples
available in the EdgeR manual, suggesting wild gene expression differences. With the exception
of sample X5A_S2, dimension 1 separates samples undergoing forced diapause (named with a "C",
mostly on the left half of the plot) from those hatching right away (named with an "A", mostly
on the right). The second dimension is also meaningful, since it almost separates populations
1, 2, and 4, which underwent random regime, from 3, 5 and 6 (regular regime). Sample X2A_S7 is
the main exception.

```{r fig.width=9, fig.height=12}
layout(mat=matrix(c(1,2,3,4,5,6,7,8,9,10,11,12), nrow=4, ncol=3), width=c(1,1,1), height=c(1,1,1,1))
for (i in 1:12) {
   plotMD(cpm(y, log=TRUE), column=i)
}
layout(mat=matrix(c(1), ncol=1, nrow=1))
```
A lot of genes seem to be differentially expressed.

#  Estimating the Biological Coefficient of Variation

```{r}
design1 <- model.matrix(~regime + diapause + regime:population)
design2 <- model.matrix(~regime + diapause + regime:population + regime:diapause)

y1 <- estimateDisp(y, design1)
y2 <- estimateDisp(y, design2)

plotBCV(y1)
plotBCV(y2)
```

# Fitting the models

```{r}
fit1 <- glmQLFit(y1, design1)
fit2 <- glmQLFit(y2, design2)

plotQLDisp(fit1)
plotQLDisp(fit2)
```

#  Contrasts
## Is the interaction term between selective regime and hatching condition necessary?

```{r}
qlf_2 <- glmQLFTest(fit2, coef=8)
genes_with_interaction_term <- topTags(qlf_2, n=dim(y)[1])$table
genes_with_interaction_term$logPValue <- -log10(genes_with_interaction_term$PValue)
ggplot(data=genes_with_interaction_term, mapping=aes(x=logFC, y=logPValue, color=FDR)) +
   geom_point() + theme_grey(base_size=20) + ylab(expression(paste("-log"[10], plain("(p value)"))))

NumGenes <- sum(genes_with_interaction_term$FDR <= 0.10)

if (NumGenes > 2) {
   fit <- fit2
} else {
   fit <- fit1
}
```

That is, if there were more than two genes with a significant interaction between the selective
regime and the hatching condition (at FDR = 0.1), then I would keep the more complex model,
including that interaction. Otherwise (which is actually the case), I keep working with the
simpler model with no interaction. Note that in both models coefficients 2 and 3 are the random
regime and the forced diapause effects.
 
