---
title: "Differential gene expression analysis with EdgeR"
output: html_document
author: "J. Ignacio Lucas Lledó"
---

```{r eval=TRUE, echo=FALSE, include=FALSE}
library(edgeR)
library(ggplot2)
```
# Reading the data

The data for this analysis was created in folder 2019-03-29. 

```{r eval=TRUE, echo=TRUE, include=TRUE}
args <- commandArgs(trailingOnly = TRUE)
if (length(args) == 0) {
   stop("At least one argument must be supplied: input file.", call.=FALSE)
} else if (length(args) == 1) {
   args[2] <- "."
}

input_file_name <- args[1]
output_dir <- args[2]
counts <- read.table(input_file_name, row.names=1, header=TRUE)
```

Three factors: population, diapause or hatching condition, and selective regime.
I prefer the immediate hatching (instead of forced diapause) and the regular or
periodic selective regime as the reference levels of the corresponding factors.

```{r}
population <- factor(c("1","1","2","2","3","3","4","4","5","5","6","6"))
NestedPop  <- factor(c("1","1","2","2","1","1","3","3","2","2","3","3"))
diapause   <- factor(rep(c("hatch","diapause"), 6))
diapause   <- relevel(diapause, "hatch")
regime     <- factor(c("random","random","random","random","regular","regular",
                 "random","random","regular","regular","regular","regular"))
regime     <- relevel(regime, "regular")
combined   <- factor(c("RandHatch","RandDiap","RandHatch","RandDiap","RegulHatch","RegulDiap",
                      "RandHatch","RandDiap","RegulHatch","RegulDiap","RegulHatch","RegulDiap"))
combined   <- relevel(combined, "RegulHatch")

y <- DGEList(counts=counts, group=regime)
y <- calcNormFactors(y, method="TMM")

barplot(y$samples$lib.size*1e-6, names=1:12, ylab="Library size (millions)")
```
It looks like the populations undergoing diapause systematically produced smaller
libraries. It may be the effect of a smaller amount of eggs available. Population 5
is the exception.


#  Filtering

The recommended filtering consists on removing genes that have such a low expression
level that do not have at least 5-10 counts on as many samples as groups there are. The
rationale is that those genes cannot have enough reads in at least one sample of each
group. However, instead of filtering on the raw counts, the edgeR manual suggests to
translate whatever minimum number of reads per library in terms of counts per million.
The reason is that the filtering should take into account differences in library sizes
between samples. It probably makes sense, because otherwise, I would be applying an
effectively higher (harsher) minimum threshold to the gene expression level in samples
with smaller libraries. That would probably bias results, since we would be enriching
the dataset in genes more highly expressed in some samples than in others.

Say that library sizes range between 250000 and 2000000. Then, applying a minimum expression
level of 20 counts per million is the only way to make sure that even the small libraries
are required to contain at least 5 counts, while at the same time setting the cutoff at the
same biologically relevant level. In any case, as long as the minimum expression
level is not enforced in all samples, the dataset retained will contain genes that do not
reach the minimum in some samples. Since counts per million are (expectedly and actually)
similarly distributed among samples, it should not be the case that any sample is enriched
in low-expression genes.

```{r}
threshold <- 5.0 / (min(y$samples$lib.size) / 1000000)
keep <- rowSums(cpm(y) > threshold) >= 4
y <- y[keep, ,keep.lib.sizes=FALSE]

plotMDS(y$counts)
```

The magnitude of variation along the axes seems very high in comparison with the examples
available in the EdgeR manual, suggesting wild gene expression differences. With the exception
of sample X5A_S2, dimension 1 separates samples undergoing forced diapause (named with a "C",
mostly on the left half of the plot) from those hatching right away (named with an "A", mostly
on the right). The second dimension is also meaningful, since it almost separates populations
1, 2, and 4, which underwent random regime, from 3, 5 and 6 (regular regime). Sample X2A_S7 is
the main exception.

```{r fig.width=9, fig.height=12}
layout(mat=matrix(c(1,2,3,4,5,6,7,8,9,10,11,12), nrow=4, ncol=3, byrow=TRUE), width=c(1,1,1), height=c(1,1,1,1))
for (i in 1:12) {
   plotMD(cpm(y, log=TRUE), column=i)
}
layout(mat=matrix(c(1), ncol=1, nrow=1))
```
A lot of genes seem to be differentially expressed. The total number of tags passing the filters is
`r dim(y)[1]`. Let's define a variable to store this value:

```{r}
NumOfTags <- dim(y)[1]
```

#  Estimating the Biological Coefficient of Variation

I need different models:

* *Model 1.* To test for an effect of the hatching condition, the most
powerful model may be this: `~population + diapause`. That is to treet samples as paired
by population. By using population, I account for variation due to selective regime.
* *Model 2.* Another simple model would ignore the blocking effect of the population
and the interaction between the two main factors: `~regime + diapause`.
* *Model 3.* If there are reasons to ignore the effect of population and the interaction
between selective regime and hatching condition, I would combine both factors in one with
four levels, following section 3.3.1 in the edgeR manual: `~0 + combined`.
* *Model 4.* Just to check for a batch effect of the population: `~regime + diapause + regime:NestedPop`.
* *Model 5.* For the sake of comparison: `~regime + diapause + regime:diapause`.
* *Model 6.* And for the sake of completeness: `~regime + diapause + regime:diapause + regime:NestedPop`.

```{r fig.width=9, fig.height=6}
design1 <- model.matrix(~population + diapause)
design2 <- model.matrix(~regime + diapause)
design3 <- model.matrix(~0 + combined)
design4 <- model.matrix(~regime + diapause + regime:NestedPop)
design5 <- model.matrix(~regime + diapause + regime:diapause)
design6 <- model.matrix(~regime + diapause + regime:diapause + regime:NestedPop)

y1 <- estimateDisp(y, design1)
y2 <- estimateDisp(y, design2)
y3 <- estimateDisp(y, design3)
y4 <- estimateDisp(y, design4)
y5 <- estimateDisp(y, design5)
y6 <- estimateDisp(y, design6)

layout(mat=matrix(c(1,2,3,4,5,6), ncol=3, nrow=2, byrow=TRUE))
plotBCV(y1, main='Model 1')
plotBCV(y2, main='Model 2')
plotBCV(y3, main='Model 3')
plotBCV(y4, main='Model 4')
plotBCV(y5, main='Model 5')
plotBCV(y6, main='Model 6')
```

# Fitting the models

```{r fig.width=9, fig.height=6}
fit1 <- glmQLFit(y1, design1)
fit2 <- glmQLFit(y2, design2)
fit3 <- glmQLFit(y3, design3)
fit4 <- glmQLFit(y4, design4)
fit5 <- glmQLFit(y5, design5)
fit6 <- glmQLFit(y6, design6)

layout(mat=matrix(c(1,2,3,4,5,6), ncol=3, nrow=2, byrow=TRUE))
plotQLDisp(fit1, main='Model 1')
plotQLDisp(fit2, main='Model 2')
plotQLDisp(fit3, main='Model 3')
plotQLDisp(fit4, main='Model 4')
plotQLDisp(fit5, main='Model 5')
plotQLDisp(fit6, main='Model 6')

layout(mat=matrix(c(1,2,3,4,5,6), ncol=3, nrow=2, byrow=TRUE))
gof(fit1, plot=TRUE, main='Model 1')
gof(fit2, plot=TRUE, main='Model 2')
gof(fit3, plot=TRUE, main='Model 3')
gof(fit4, plot=TRUE, main='Model 4')
gof(fit5, plot=TRUE, main='Model 5')
gof(fit6, plot=TRUE, main='Model 6')
```

# Genes affected by hatching condition
## Model 1

Model 1 is expected to be the most sensitive to detect genes the expression of which is
affected by enforcing diapause, after accounting for the variation among populations (which
includes variation due to selective regime).

```{r}
qlf1 <- glmQLFTest(fit1)
topTags(qlf1)
z1 <- topTags(qlf1, n=NumOfTags)$table
z1$logPValue <- -log10(z1$PValue)
ggplot(data=z1, mapping=aes(x=logFC, y=logPValue, color=FDR)) +
   geom_point() + theme_grey() + ylab(expression(paste("-log"[10], plain("(p value)")))) +
   ggtitle('Effect of enforcing diapause, model 1')
```

I would say the two first genes (XLOC_052427 and XLOC_045087) are significantly downregulated
when enforcing diapause, with respect to the immediate hatching condition.

## Model 2

Model 2 is very simple, and may include false positives, because it does not account for any
interaction, nor for among-population variance.

```{r}
qlf2 <- glmQLFTest(fit2, coef=3)
topTags(qlf2)
z1 <- topTags(qlf2, n=NumOfTags)$table
z1$logPValue <- -log10(z1$PValue)
ggplot(data=z1, mapping=aes(x=logFC, y=logPValue, color=FDR)) +
   geom_point() + theme_grey() + ylab(expression(paste("-log"[10], plain("(p value)")))) +
   ggtitle('Effect of enforcing diapause, model 2')
```

According to model 2, there are `r dim(topTags(qlf2, n=NumOfTags, p.value=0.10))[1]` tags
significantly up- or downregulated by hatching condition, at FDR=0.1.

## Model 3

In model 3, the effect of diapause may be understood in two different ways. First, we may want
to know what genes have an average expression level different between hatching conditions. This
is different from testing the effect of hatching condition twice: first among samples in a regular
environment and then among those in the stochastic environment. Then, I could select the genes
with the same significant tendency in both selective regimes. I will limit this analysis to the
first option, comparing only average gene expression level.

```{r}
qlf3 <- glmQLFTest(fit3, contrast=c(-0.5, 0.5, -0.5, 0.5))
topTags(qlf3)
z1 <- topTags(qlf3, n=NumOfTags)$table
z1$logPValue <- -log10(z1$PValue)
ggplot(data=z1, mapping=aes(x=logFC, y=logPValue, color=FDR)) +
   geom_point() + theme_grey() + ylab(expression(paste("-log"[10], plain("(p value)")))) +
   ggtitle('Effect of enforcing diapause, model 3')
```

According to model 3, there are `r dim(topTags(qlf3, n=NumOfTags, p.value=0.10))[1]` tags
significantly up- or downregulated by hatching condition at FDR = 0.1.

## Model 4

```{r}
qlf4 <- glmQLFTest(fit4, coef=3)
topTags(qlf4)
z1 <- topTags(qlf4, n=NumOfTags)$table
z1$logPValue <- -log10(z1$PValue)
ggplot(data=z1, mapping=aes(x=logFC, y=logPValue, color=FDR)) +
   geom_point() + theme_grey() + ylab(expression(paste("-log"[10], plain("(p value)")))) +
   ggtitle('Effect of enforcing diapause, model 4')
```

According to model 4, there are `r dim(topTags(qlf4, n=NumOfTags, p.value=0.10))[1]` tags
significantly up- or downregulated by hatching condition at FDR = 0.1.

## Model 5

```{r}
qlf5 <- glmQLFTest(fit5, coef=3)
topTags(qlf5)
z1 <- topTags(qlf5, n=NumOfTags)$table
z1$logPValue <- -log10(z1$PValue)
ggplot(data=z1, mapping=aes(x=logFC, y=logPValue, color=FDR)) +
   geom_point() + theme_grey() + ylab(expression(paste("-log"[10], plain("(p value)")))) +
   ggtitle('Effect of enforcing diapause, model 5')
```

According to model 5, there are `r dim(topTags(qlf5, n=NumOfTags, p.value=0.10))[1]` tags
significantly up- or downregulated by hatching condition at FDR = 0.1.

## Model 6

```{r}
qlf6 <- glmQLFTest(fit6, coef=3)
topTags(qlf6)
z1 <- topTags(qlf6, n=NumOfTags)$table
z1$logPValue <- -log10(z1$PValue)
ggplot(data=z1, mapping=aes(x=logFC, y=logPValue, color=FDR)) +
   geom_point() + theme_grey() + ylab(expression(paste("-log"[10], plain("(p value)")))) +
   ggtitle('Effect of enforcing diapause, model 6')
```

According to model 6, there are `r dim(topTags(qlf6, n=NumOfTags, p.value=0.10))[1]` tags
significantly up- or downregulated by hatching condition at FDR = 0.10. And this is how
the results from models 2, 4, 5 and 6 match:

```{r}
significant <- cbind(model2 = topTags(qlf2, n=NumOfTags)$table$FDR <= 0.1, 
                     model3 = topTags(qlf3, n=NumOfTags)$table$FDR <= 0.1,
                     model4 = topTags(qlf4, n=NumOfTags)$table$FDR <= 0.1,
                     model5 = topTags(qlf5, n=NumOfTags)$table$FDR <= 0.1,
                     model6 = topTags(qlf6, n=NumOfTags)$table$FDR <= 0.1)
vennDiagram(vennCounts(significant))
```

# Interaction between selective regime and hatching condition
## Model 6

```{r}
qlf6 <- glmQLFTest(fit6, coef=4)
topTags(qlf6)
z1 <- topTags(qlf6, n=NumOfTags)$table
z1$logPValue <- -log10(z1$PValue)
ggplot(data=z1, mapping=aes(x=logFC, y=logPValue, color=FDR)) + 
   geom_point() + theme_grey() + ylab(expression(paste("-log"[10], plain("(p value)")))) +
   ggtitle('Interaction regime-hatching')
```

According to model 6, there are `r dim(topTags(qlf6, n=NumOfTags, p.value=0.25))[1]` tags
with a significant interaction term between selective regime and hatching condition at
FDR = 0.25.

## Model 5

```{r}
qlf5 <- glmQLFTest(fit5, coef=4)
topTags(qlf5)
z1 <- topTags(qlf5, n=NumOfTags)$table
z1$logPValue <- -log10(z1$PValue)
ggplot(data=z1, mapping=aes(x=logFC, y=logPValue, color=FDR)) +
   geom_point() + theme_grey() + ylab(expression(paste("-log"[10], plain("p value)")))) + 
   ggtitle('Interaction regime-hatching in model 5')
```

According to model 5, there are `r dim(topTags(qlf5, n=NumOfTags, p.value=0.25))[1]` tags
with a significant interaction term between selective regime and hatching condition at 
FDR = 0.25. And this is how models 5 and 6 agree on what genes have such significant interaction
term:

```{r}
significant <- cbind(model5 = topTags(qlf5, n=NumOfTags)$table$FDR <= 0.25,
                     model6 = topTags(qlf6, n=NumOfTags)$table$FDR <= 0.25)
vennDiagram(vennCounts(significant))
```

# Interaction between selective regime and population

When testing for significant coefficients showing the effect of the nested population, there
are not one but 4 possible coefficients, each with a different fold change. For the volcano
plot, I want to use for each gene the fold change that is most extreme (either positive or
negative).

## Model 6

```{r}
qlf6  <- glmQLFTest(fit6, coef=5:8)
topTags(qlf6)
z1 <- topTags(qlf6, n=NumOfTags)$table
z1$logPValue <- -log10(z1$PValue)
z1$logFC <- z1[cbind(1:NumOfTags, apply(abs(z1[,1:4]), 1, which.max))]
ggplot(data=z1, mapping=aes(x=logFC, y=logPValue, color=FDR)) +
   geom_point() + theme_grey() + ylab(expression(paste("-log"[10], plain("p value)")))) +
   ggtitle('Interaction regime-population in complete model')
```

According to model 6, there are `r dim(topTags(qlf6, n=NumOfTags, p.value=0.01))[1]` tags
with a significant interaction term between selective regime and population, at FDR = 0.01.

## Model 4

```{r}
qlf4 <- glmQLFTest(fit4, coef=4:7)
topTags(qlf4)
z1 <- topTags(qlf4, n=NumOfTags)$table
z1$logPValue <- -log10(z1$PValue)
z1$logFC <- z1[cbind(1:NumOfTags, apply(abs(z1[,1:4]), 1, which.max))]
ggplot(data=z1, mapping=aes(x=logFC, y=logPValue, color=FDR)) +
   geom_point() + theme_grey() + ylab(expression(paste("-log"[10], plain("p value)")))) + 
   ggtitle('Interaction regime-population in model 4')
```

According to model 4, there are `r dim(topTags(qlf4, n=NumOfTags, p.value=0.01))[1]` tags
with a significant interaction term between selective regimew and population, at FDR = 0.01.
And this is how the genes identified as having such significant coefficients match between
results of models 4 and 6:

```{r}
significant <- cbind(model4 = topTags(qlf4, n=NumOfTags)$table$FDR <= 0.01,
                     model6 = topTags(qlf6, n=NumOfTags)$table$FDR <= 0.01)
vennDiagram(vennCounts(significant))
```

# Main effect of the selective regime

## Model 2

```{r}
qlf2  <- glmQLFTest(fit2, coef=2)
topTags(qlf2)
z1 <- topTags(qlf2, n=NumOfTags)$table
z1$logPValue <- -log10(z1$PValue)
ggplot(data=z1, mapping=aes(x=logFC, y=logPValue, color=FDR)) +
   geom_point() + theme_grey() + ylab(expression(paste("-log"[10], plain("p value)")))) +
   ggtitle('Main effect of selective regime')
```

According to model 2, there are `r dim(topTags(qlf2,n=NumOfTags,p.value=0.01))[1]` tags
significantly up- or downregulated by selective regime at FDR=0.01.

## Model 3

Using model 3 I can look for genes affected by selective regime in two ways. First, I could
look for a significant change in average gene expression among samples in the same regime.
Alternatively, I could separately look for genes affected by the selective regime among samples
in the same hatching condition, and then compare the lists of significant genes. According to
Edu, the latter makes more sense biologically, because we expect the selective regime to affect
different sets of genes depending on whether eggs are allowed to hatch or *forced* to undergo
diapause.

### Effect of selective regime when eggs are allowed to hatch immediately

```{r}
qlf3a  <- glmQLFTest(fit3, contrast=c(-1, 0, 1, 0))
topTags(qlf3a)
z1 <- topTags(qlf3a, n=NumOfTags)$table
z1$logPValue <- -log10(z1$PValue)
ggplot(data=z1, mapping=aes(x=logFC, y=logPValue, color=FDR)) +
   geom_point() + theme_grey() + ylab(expression(paste("-log"[10], plain("p value)")))) +
   ggtitle('Effect of selective regime, when hatching')
```
According to model 3, contrast *a*, there are `r dim(topTags(qlf3a, n=NumOfTags, p.value=0.05))[1]` tags
differentially expressed between selective regimes among samples allowed to hatch immediately, at FDR=0.05.

### Effect of selective regime when eggs are forced to undergo diapause

```{r}
qlf3b  <- glmQLFTest(fit3, contrast=c(0, 1, 0, -1))
topTags(qlf3b)
z1 <- topTags(qlf3b, n=NumOfTags)$table
z1$logPValue <- -log10(z1$PValue)
ggplot(data=z1, mapping=aes(x=logFC, y=logPValue, color=FDR)) +
   geom_point() + theme_grey() + ylab(expression(paste("-log"[10], plain("p value)")))) +
   ggtitle('Effect of selective regime, when hatching')
```

According to model 3, contrast *b*, there are `r dim(topTags(qlf3b, n=NumOfTags, p.value=0.05))[1]` tags
differentially expressed between selective regimes among samples undergoing diapause, at FDR=0.05. There is
a reasonable correlation between the fold change in the two hatching conditions:

```{r}
z2 <- merge(topTags(qlf3a, n=NumOfTags)$table,
            topTags(qlf3b, n=NumOfTags)$table,
            by='row.names')
ggplot(data=z2, mapping=aes(x=logFC.x, y=logFC.y, color=FDR.x)) +
   geom_point() + geom_smooth(method='lm') + xlab('logFC hatching') + ylab('logFC diapause')
```

### Average effect of selective regime

```{r}
qlf3c  <- glmQLFTest(fit3, contrast=c(-0.5, 0.5, 0.5, -0.5))
topTags(qlf3c)
z1 <- topTags(qlf3c, n=NumOfTags)$table
z1$logPValue <- -log10(z1$PValue)
ggplot(data=z1, mapping=aes(x=logFC, y=logPValue, color=FDR)) +
   geom_point() + theme_grey() + ylab(expression(paste("-log"[10], plain("p value)")))) +
   ggtitle('Effect of selective regime, when hatching')
```

According to model 3, contrast *c*, there are `r dim(topTags(qlf3b, n=NumOfTags, p.value=0.05))[1]` tags
with a significant change (across selective regimes) of average (across hatching conditions) expression
levels, at FDR = 0.05. 

## Model 4

```{r}
qlf4  <- glmQLFTest(fit4, coef=2)
topTags(qlf4)
z1 <- topTags(qlf4, n=NumOfTags)$table
z1$logPValue <- -log10(z1$PValue)
ggplot(data=z1, mapping=aes(x=logFC, y=logPValue, color=FDR)) +
   geom_point() + theme_grey() + ylab(expression(paste("-log"[10], plain("p value)")))) +
   ggtitle('Main effect of selective regime, model 4')
```

According to model 4, there are `r dim(topTags(qlf4,n=NumOfTags,p.value=0.25))[1]` tags
significantly up- or downregulated by selective regime at FDR=0.25.

## Model 5

```{r}
qlf5  <- glmQLFTest(fit5, coef=2)
topTags(qlf5)
z1 <- topTags(qlf5, n=NumOfTags)$table
z1$logPValue <- -log10(z1$PValue)
ggplot(data=z1, mapping=aes(x=logFC, y=logPValue, color=FDR)) +
   geom_point() + theme_grey() + ylab(expression(paste("-log"[10], plain("p value)")))) +
   ggtitle('Main effect of selective regime, model 5')
```

According to model 5, there are `r dim(topTags(qlf5,n=NumOfTags,p.value=0.05))[1]` tags
significantly up- or downregulated by selective regime at FDR=0.25.

## Model 6

```{r}
qlf6  <- glmQLFTest(fit6, coef=2)
topTags(qlf6)
z1 <- topTags(qlf6, n=NumOfTags)$table
z1$logPValue <- -log10(z1$PValue)
ggplot(data=z1, mapping=aes(x=logFC, y=logPValue, color=FDR)) +
   geom_point() + theme_grey() + ylab(expression(paste("-log"[10], plain("p value)")))) +
   ggtitle('Main effect of selective regime, model 6')
```

According to model 6, there are `r dim(topTags(qlf6,n=NumOfTags,p.value=0.25))[1]` tags
significantly up- or downregulated by selective regime at FDR=0.25. And this is how the
different models agree on what genes are differentially expressed in different selective
regimes:

```{r}
significant <- cbind(model2 = topTags(qlf2,  n=NumOfTags)$table$FDR <= 0.01,
                     model3 = topTags(qlf3a, n=NumOfTags)$table$FDR <= 0.05,
                     model4 = topTags(qlf4,  n=NumOfTags)$table$FDR <= 0.25,
                     model5 = topTags(qlf5,  n=NumOfTags)$table$FDR <= 0.05,
                     model6 = topTags(qlf6,  n=NumOfTags)$table$FDR <= 0.25)
vennDiagram(vennCounts(significant))
```

Note that I am using a different FDR for each model, to obtain more or less comparable
and positive numbers.
