---
title: "Comparison of variation of expression level between regimes"
author: "J. Ignacio Lucas Lledó"
date: "28/1/2020"
output: html_document
bibliography: variation.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

```{r setup2, message=FALSE}
library(variancePartition)
library(ggplot2)
library(tidyr)
library(reactable)
library(BiocParallel)
library(edgeR)

EXPRESSION <- '../2020-01-08/genes.RData'   # variance partition and the mixed models fitted with dream()
COUNTS <- '../2019-03-29/genes.PostCount.txt'
NUM_THREADS <- 2
```

One obvious expectation that I have not tested yet is that resting eggs under an unpredictable
environment should also have a more *random*, noiser, or unpredictable gene expression. That is,
gene expression levels should vary more among eggs laid in unpredictable environments. To put it
another way, in perfectly regular environments, eggs are expected to resemble more each other in
terms of gene expression patters. Gene expression should be less variable among eggs in regular
than in unpredictable environments.

We did not measure expression levels in single cells. But the variance of expression levels among
individual cells contributes to the observed variance among samples, from which the mean-variance
relationshps were estimated. The `variancePartition` package does not include a function to plot
the estimated mean-variance relationship. That relationship is used to impute the variance of
every single mesurement of expression level in every sample. The inverse of that imputed variance
is stored in the `EList` object created with the `voomWithDreamWeights()` function, that I
named `vobjDream`. 

```{r variance, message=FALSE}
load(EXPRESION)
colnames(vobjDream$E)       <- paste('expression', row.names(vobjDream$targets), sep='_')
colnames(vobjDream$weights) <- paste('variance', row.names(vobjDream$targets), sep='_')
meanVar <- as.data.frame(cbind(vobjDream$E, 1/vobjDream$weights))
meanVar$gene <- row.names(meanVar)    # column 25
meanVar <- pivot_longer(meanVar, cols=1:24, names_to=c('what','population','hatching'),
                        names_pattern='(expression|variance)_(X[1-6])([AC])', values_to='value')
meanVar$regime <- factor('regular', levels=c('regular', 'random'))
meanVar[meanVar$population %in% c('X1', 'X2', 'X4'), 'regime'] <- 'random'
meanVar <- pivot_wider(meanVar, names_from='what', values_from='value')
ggplot(data=meanVar, mapping=aes(x=expression, y=variance, color=regime)) +
  geom_smooth() + xlab('log2 of counts per million')
```

Yes, eggs have more variable expression levels in unpredictable environments. The difference
is not that big, actually. It makes me think that heteroscedasticity may not be properly
taken into account. Even though variance is modelled at the observation level, variances of
individual observations are predicted from the mean-variance trend estimated from gene-wise
mean expression levels and residual standard deviations (from gene-wise linear fits)
[@Law2014]. That is, the variance predictions are informed by the expression level expected
under a gene-wise linear model. Such models do include the effects of all variables. However,
I don't think the systematic effect of the selective regime in gene expression variance is
taken into account.

# Separate estimation of variances for each selective regime

```{r separate}
counts <- as.matrix(read.table(COUNTS, row.names=1, header=TRUE))
threshold <- 5 / (min(colSums(counts)) / 1000000)
keep <- rowSums(cpm(counts) > threshold) >= 4

# samples:                 1A 1C 2A 2C 3A 3C 4A 4C 5A 5C 6A 6C
countsRandom  <- counts[keep,c( 1, 2, 3, 4,       7, 8            )]
countsRegular <- counts[keep,c(             5, 6,       9,10,11,12)]
geneExprRandom  <- DGEList(countsRandom)
geneExprRandom  <- calcNormFactors(geneExprRandom)
geneExprRegular <- DGEList(countsRegular)
geneExprRegular <- calcNormFactors(geneExprRegular)

metadataRandom  <- data.frame(row.names  = c("X1A_S8", "X1C_S1", "X2A_S7", "X2C_S5", "X4A_S6", "X4C_S12"),
                              population = factor(as.character(c(1,1,2,2,3,3)), levels=c('1','2','3')),
                              treatment  = c(0, 1, 0, 1, 0, 1),
                              treatmentChar = factor(c('hatch','diapause','hatch','diapause','hatch','diapause')))
metadataRegular <- data.frame(row.names  = c("X3A_S9", "X3C_S11", "X5A_S2", "X5C_S4", "X6A_S3", "X6C_S10"),
                              population = factor(as.character(c(1,1,2,2,3,3)), levels=c('1','2','3')),
                              treatment  = c(0, 1, 0, 1, 0, 1),
                              treatmentChar = factor(c('hatch','diapause','hatch','diapause','hatch','diapause')))

form <- ~ treatment + (1|population)

param = SnowParam(NUM_THREADS, "SOCK")
register(param)

## Weights
weightsRandom  <- voomWithDreamWeights(geneExprRandom,  form, metadataRandom)
weightsRegular <- voomWithDreamWeights(geneExprRegular, form, metadataRegular)
```

```{r plot, message=FALSE}
E <- cbind(weightsRandom$E, weightsRegular$E)[,c(1,2,3,4,7,8,5,6,9,10,11,12)]
W <- cbind(weightsRandom$weights, weightsRegular$weights)[,c(1,2,3,4,7,8,5,6,9,10,11,12)]
colnames(E) <- paste('E', row.names(vobjDream$targets), sep='_')
colnames(W) <- paste('V', row.names(vobjDream$targets), sep='_')
meanVarSep  <- as.data.frame(cbind(E, 1/W))
meanVarSep$gene <- row.names(E)    # column 25
meanVarSep <- pivot_longer(meanVarSep, cols=1:24, names_to=c('what','population','hatching'),
                           names_pattern='([EV])_(X[1-6])([AC])', values_to='value')
meanVarSep$regime <- factor('regular', levels=c('regular', 'random'))
meanVarSep[meanVarSep$population %in% c('X1', 'X2', 'X4'), 'regime'] <- 'random'
meanVarSep <- pivot_wider(meanVarSep, names_from='what', values_from='value')
ggplot(data=meanVarSep, mapping=aes(x=E, y=V, color=regime)) +
  geom_smooth() + xlab('log2 of counts per million')
```

# References