---
title: "Variation"
author: "J. Ignacio Lucas Lledó"
date: "3/7/2020"
output: html_document
bibliography: variation.bib
---

```{r setup, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(DESeq2)
library(ggplot2)
library(vegan)
library(PoiClaClu)

COUNTS <- '../2019-03-29/genes.PostCount.txt'
```

The counts of reads overlapping a gene in a sample of dormant eggs is determined
by the sequencing effort (library size) and the unobserved true proportion of
that gene's transcripts in the set of dormant eggs. In turn, that unobserved
proportion is the average of all individual cells in all eggs in the sample. The
variance of the sample-wise unobserved proportion of a gene's transcripts can be
thought of as the variance of the average of all cell-wise unobserved proportions.
That is, the biological variation between two replicates is made up of the variation
in their cells, of course. The sample-wise variance is the variance of an average,
which is proportional to the variances and covariances of the cell-wise values.

When performing RNA-seq experiments on anything other than single cells, positive
correlations are implicitly assumed: two cells may be expected to have correlated
expressions of a gene for being part of the same tissue, the same organism, or for
being under similar conditions. The experiment would not make sense if samples were
composed of cells with perfect negative correlations, in a way that gene expression
changes in, say, half the cells were compensated by changes in the opposite direction
in the other half of cells. In such an scenario, any two samples would look alike,
despite wide variation at the cell level. Actually, even at the cell level they would
look alike, unless cells could be enumerated or classified to be compared between samples.

Thus, it is perfectly reasonable to expect that if an environment makes the expression
of a gene more variable among cells (or eggs) it will also make the sample-wise
proportion of that gene's transcript more variable among samples. 

Packages like `edgeR` and `DESeq2` that test for differences of gene expression
between conditions need to estimate the variance of read counts
from few replicates. They share information among genes to compensate for the
scarcity of replicates, assuming that genes with similar expression levels must
have similar amounts of variation among replicates. In order to obtain estimates
of variance for individual values of read counts, they use the observed relationship
between dispersion and mean value. In [@Anders2010] it is quite clear that DESeq2 estimates
variance of individual values separately for the different groups of samples.
It is less clear that `edgeR` allows for a different variance in
different conditions. That would explain why the mean-variance relationship changes
so much when estimating variance separately for the two regimes when using
`edgeR` (see `2020-01-28`). Here I will use DESeq2 to compare the the variance
of gene expression between conditions. I found [This](https://bioconductor.org/packages/release/workflows/vignettes/rnaseqGene/inst/doc/rnaseqGene.html#sample-distances)
interesting tutorial.

```{r data, message=FALSE}
counts <- as.matrix(read.table(COUNTS, row.names=1, header=TRUE))
samples <- data.frame(row.names = c("X1A_S8", "X1C_S1", "X2A_S7", "X2C_S5",
                                    "X3A_S9", "X3C_S11", "X4A_S6", "X4C_S12",
                                    "X5A_S2", "X5C_S4", "X6A_S3", "X6C_S10"),
                      regime = c("unpredictable", "unpredictable", "unpredictable",
                                 "unpredictable", "predictable", "predictable",
                                 "unpredictable", "unpredictable", "predictable",
                                 "predictable", "predictable", "predictable"),
                      population = c("1", "1", "2", "2", "3", "3",
                                     "4", "4", "5", "5", "6", "6"),
                      treatment = rep(c("hatch", "diapause"), 6))
stopifnot(all.equal(rownames(samples), colnames(counts)))
dds <- DESeqDataSetFromMatrix(countData = round(counts), colData = samples, design = ~ regime)
keep <- rowSums(counts(dds)) > 1
dds <- dds[keep,]
nrow(dds)
```

# Estimating dispersions with DESeq2

```{r dispersion, message=FALSE}
dds <- estimateSizeFactors(dds)
dds <- estimateDispersions(dds, fitType = 'parametric')
```

The `estimateDispersions()` function estimates dispersions by maximization of
a Cox Reid-adjusted profile likelihood, which is a penalized likelihood that
accounts for the usage of data to estimate both the dispersion and the coefficients
of the GLM of the expression levels themselves. That is, the design formula is
used in the estimation of gene-wise dispersions. But the estimated dispersions
are not separate for separate conditions. The model assumes a gene's expression
has the same coefficient of variation in all sample groups. This is described in
[@McCarthy2012]. Thus, it does not help compare the dispersions among sample
groups.

```{r split, message=FALSE}
is.random   <- samples$regime == 'unpredictable'
is.regular  <- samples$regime == 'predictable'
dds.random  <- DESeqDataSetFromMatrix(countData = round(counts[, is.random]),
                                      colData = samples[is.random,], design = ~ 1)
dds.regular <- DESeqDataSetFromMatrix(countData = round(counts[, is.regular]),
                                      colData = samples[is.regular,], design = ~ 1)
keep <- rowSums(counts(dds.random)) > 1 & rowSums(counts(dds.regular)) > 1
dds.random  <- dds.random[keep,]
dds.regular <- dds.regular[keep,]
dds.random  <- estimateSizeFactors(dds.random)
dds.regular <- estimateSizeFactors(dds.regular)
dds.random  <- estimateDispersions(dds.random,  fitType = 'parametric')
dds.regular <- estimateDispersions(dds.regular, fitType = 'parametric')

MeanDisp <- rbind(data.frame(Regime = factor('unpredictable',
                                             levels=c('predictable','unpredictable')),
                             Dispersion = dispersions(dds.random),
                             MeanCount  = mcols(dds.random)$baseMean),
                  data.frame(Regime = factor('predictable',
                                             levels=c('predictable','unpredictable')),
                             Dispersion = dispersions(dds.regular),
                             MeanCount = mcols(dds.regular)$baseMean))

ggplot(data = MeanDisp, mapping = aes(x = MeanCount, y = Dispersion, color = Regime)) +
   geom_smooth() + scale_x_log10() + xlab('Mean count') + ylab('Dispersion')
ggsave('meandisp.png')
```

This is qualitatively similar to the comparison made in `2020-01-28` using
`edgeR`: random environments make low-expressed genes be more variably
expressed among samples (and presumably among eggs).

# Homogeneity test

I would like a more formal comparison of dispersion between regimes. Another
approach is to reduce the dimensionality of the counts matrix and compare the
samples using any distance function. And then test for homogeneity between
regimes [@Anderson2006] using the `vegan` package [@Oksanen2019]. The distance
measure I want to use is the Poisson distance implemented in the `PoiClaClu`
package, by @Witten2011.

```{r homogeneity, message=FALSE}
poisd <- PoissonDistance(t(counts(dds)))
anova(betadisper(poisd$dd, samples$regime))

permutest(betadisper(poisd$dd, samples$regime))

plot(betadisper(poisd$dd, samples$regime), main='Poisson distances',
     sub='')
png(filename = 'poisson.png')
  plot(betadisper(poisd$dd, samples$regime), sub='', main='')
dev.off()
```

Using the formal test, the difference in among-sample dispersion between
regimes is not significant, even though the average distance to the
centroid is lower in the regular than in the random environment. That was
using the Poisson distance, which is appropriate for count data. Let's
try with a transformation of the count data that stabilizes the variance.

```{r transformations}
rld <- rlog(dds, blind=TRUE)
sampleDist <- dist(t(assay(rld)))
permutest(betadisper(sampleDist, samples$regime))
plot(betadisper(sampleDist, samples$regime), main='rlog transformed counts, Euclidean distances')
```

Hmmm, the samples from the *random* environment make up a Star Destroyer...

# Conclusions

The unpredictable environment makes the expression of lowly expressed genes
more variable, but it does not affect the genes expressed at high levels
in the same way. As a result, the dispersion in the overall composition of
RNA species among samples is not significantly higher among samples from the
unpredictable environment.

# Session information

```{r sessionInfo}
sessionInfo()
```

# References