---
title: "Variance partition of isoform expression in a split-plot design"
author: "J. Ignacio Lucas Lledó"
date: "9/1/2020"
output: html_document
---

# Introduction
The EdgeR package has the advantage of modelling count data, instead of normalized expression
levels. It offers some capacity to fit complex models. But it does not include random effects.
The experiment was a split-plot design, which requires a mixed model with the random effect of
the original population of origin in order to properly account for the correlation of expression
levels in sub-populations of the same origin.

The `variancePartition` package fits mixed models to the expression data of every gene in the
dataset. It also includes the `dream()` function to test for differential expression under this
framework. It does require the expression level to be normalized, though.

```{r setup}
library(variancePartition)
library(edgeR)
library(BiocParallel)
COUNTS_FILE <- '../2019-03-29/isoforms.PostCount.txt'
NUM_THREADS <- 40  # Number of cores for parallelization.
TAG <- 'isoforms'
```

# Variance partition
When fitting the model to every gene's expression profile, information from all genes contribute to
every fit. The shared information helps determine the dispersion of every measure. Counts are heteroscedastic:
their variance depend on their mean. The exact mean-variance relationship may be complex, but it can
be learned from the data. The `limma` package uses the `voom()` function to 1) transform count data to
log2-counts per million, and to estimate the mean-variance relationship. Once that relationship is estimated,
every expression level is assigned a *weight*: the inverse of its predicted variance. This weights account for
the heteroscedasticity and are used afterwards in fitting the model and in testing for differential expression.

Curiously, though, a model needs to be fit first, for every gene. The residual standard deviation and the
average gene expression level of that fit are one point in the scatterplot where the mean-variance relationship
becomes apparent. When fitting these gene-wise linear models, `voom()` does not consider random errors, but
`voomWithDreamWeights()` does. 

```{r data, cache=TRUE}
counts <- as.matrix(read.table(COUNTS_FILE, row.names=1, header=TRUE))

## Filtering
# An isoform is expressed if it has at least 5 reads, which in terms of CPM in the smallest
# library is equal to `threshold`. Only genes expressed in at least 4 samples are retained.
threshold <- 5 / (min(colSums(counts)) / 1000000)
keep <- rowSums(cpm(counts) > threshold) >= 4
geneExpr <- DGEList(counts[keep,])
geneExpr <- calcNormFactors(geneExpr)

## Metadata
metadata <- data.frame(row.names = c("X1A_S8", "X1C_S1", "X2A_S7", "X2C_S5",
                                  "X3A_S9", "X3C_S11", "X4A_S6", "X4C_S12",
                                  "X5A_S2", "X5C_S4", "X6A_S3", "X6C_S10"),
                       regime = c(1, 1, 1, 1, 0, 0 , 1, 1, 0, 0, 0, 0),
                       regimeChar = c("random", "random", "random", "random",
                                  "regular", "regular", "random", "random",
                                  "regular", "regular", "regular", "regular"),
                       population = c("1", "1", "2", "2", "3", "3",
                                      "4", "4", "5", "5", "6", "6"),
                       treatment = rep(c(0, 1), 6),
                       treatmentChar = rep(c("hatch", "diapause"), 6))
form <- ~ regime + treatment + (1|population)
```

The `variancePartition` package requires all categorical variables to be random effects. This is a way to prevent
bias when estimating the variance component of categorical variables with many levels. However, the
`dream()` function seems to require the variable to be tested be a fixed effect. To get out of this catch22,
I dummy-code regime and treatment as numeric variables with values 0 and 1.

```{r fits, cache=TRUE, warning=FALSE, message=FALSE}
## Parallelization
param = SnowParam(NUM_THREADS, "SOCK")
register(param)

## Weights
vobjDream <- voomWithDreamWeights(geneExpr, form, metadata)

## Variance partition
varPart <- fitExtractVarPartModel(vobjDream, form, metadata)
plotVarPart(varPart)
plotPercentBars(varPart[1:20,])
```

Recall that the partition of variance shows relative amounts of variance explained by individual predictors,
irrespectively of the absolute amount of variation. Thus, isoforms for which a large fraction of their expression
variance is explained by the selective regime do not necessarily have large fold change values. Let's take a
look at the stratified expression level of the isoform with largest fraction of variance explained by regime:

```{r stratified}
i <- which.max(varPart$regime)

# Expression level is log2 counts per million.
GE <- data.frame(Expression = vobjDream$E[i,], Regime = metadata$regimeChar)
plotStratify(Expression ~ Regime, GE, main = rownames(vobjDream$E)[i])
```

```{r dream, cache=TRUE}
fitmm = dream( vobjDream, form, metadata )
```

# Session Info

```{r sessioninfo}
save(vobjDream, varPart, fitmm, file = paste(TAG, '.RData', sep=''))
sessionInfo()
```

# References
