---
title: "Cell functions affected by environmental unpredictability"
author: "J. Ignacio Lucas Lledó"
date: "23/1/2020"
output: html_document
bibliography: interpretation.bib
---

```{r hide, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

On `2020-01-14` I run the analysis of Gene Ontology terms. Here I want to focus on part of those results and make some findings
clearer. I will use results only from gene annotation, and not from annotation of transcripts, which is a somewhat noiser dataset.
I will not look at the functions affected by hatching condition, since it only affects a couple of genes, to be inspected later.
Among all cellular functions significantly enriched with genes differentially expressed among selective regimes, I will focus
on those that are most significant, and detailed enough. For example, I skip terms like *transmembrane transport*, *proteolysis*
or *protein phosphorylation*, because they are too general. Sections below are named for the functional categories that I think
are worth discussing. I will try to contextualize the results. My entry point to the literature is [GarciaRoger2019].

I need the results of the differential expression analysis (`2020-01-08`) to identify the genes annotated with the significant
GO terms, and to assess the direction of gene expression change between selective regimes.

```{r setup, message=FALSE}
library(variancePartition)
library(GO.db)
library(ggplot2)
library(tidyr)
ANNOTATION <- '../2019-07-26/genes/annotation.txt'
EXPRESSION <- '../2020-01-08/genes.RData'   # Includes the variance partition and the mixed models fitted with dream()
```

The annotation table includes only the most specific terms that can be assigned to a
gene. The more general, ancestor terms are assumed, of course, but not explicitly
mentioned in the annotation. For the purpose of retrieving all genes annotated to a
term, I need to expand the annotation, to make ancestor terms explicit. It took me a
while to realize that I can do this with the `unlist()` function:

```{r data}
annotation <- read.table(ANNOTATION, col.names = c('gene', 'GOterms'), colClasses = c('character', 'character'))
head(annotation)
annotation.list <- strsplit(annotation$GOterms, split='|', fixed=TRUE)
names(annotation.list) <- annotation$gene
# append can only join two vectors at a time. It works with lists!
allAncestors <- append(as.list(GOBPANCESTOR),
                       append(as.list(GOMFANCESTOR),
                              as.list(GOCCANCESTOR)))

fullAnnotation <- lapply(annotation.list,
                         FUN = function(x){
                           unique(append(x,
                                         unlist(allAncestors[x], use.names=FALSE)))
                         })
load(EXPRESSION)
ls()
head(varPart)
```

# Biological processes
## Nucleotide-excision repair
This is term **GO:0006289**. It is among the most significantly enriched terms, with only 13 genes annotated.

```{r NER}
goterm <- 'GO:0006289'
genes <- names(fullAnnotation[grep(goterm, fullAnnotation, fixed=TRUE)])
# Not all genes annotated have been used in expression analysis
genes <- genes[genes %in% row.names(varPart)]
plotPercentBars(varPart[genes,])

topTable(fitmm, coef='regime', number=length(fitmm$F))[genes,]

z <- as.data.frame(vobjDream$E[genes,])
z$gene <- row.names(z)
GE <- pivot_longer(z, cols=c(1:12), names_to='sample', values_to='expression')
GE$regime <- factor(NA, levels=c('regular', 'random'))
GE[GE$sample %in% c("X1A_S8", "X1C_S1", "X2A_S7", "X2C_S5", "X4A_S6", "X4C_S12"), 'regime'] <- 'random'
GE[GE$sample %in% c("X3A_S9", "X3C_S11", "X5A_S2", "X5C_S4", "X6A_S3", "X6C_S10"), 'regime'] <- 'regular'
ggplot(data=GE, mapping=aes(x=regime, y=expression, color=regime)) + geom_boxplot() + facet_wrap(~gene) + ggtitle(Term(GOTERM[goterm]))
```

## G protein-coupled receptor signaing pathway, and signal transduction

```{r Gprotein}
goterm <- 'GO:0007186'
genes <- names(fullAnnotation[grep(goterm, fullAnnotation, fixed=TRUE)])
genes <- genes[genes %in% row.names(varPart)]
# I subset varPart and at the same time order the selected rows by 'regime', and the columns in the order desired.
plotPercentBars(varPart[genes[order(-varPart[genes, 'regime'])], c('regime', 'treatment', 'population', 'Residuals')])

z <- as.data.frame(vobjDream$E[genes,])
z$gene <- row.names(z)
GE <- pivot_longer(z, cols=c(1:12), names_to='sample', values_to='expression')
GE$regime <- factor(NA, levels=c('regular', 'random'))
GE[GE$sample %in% c("X1A_S8", "X1C_S1", "X2A_S7", "X2C_S5", "X4A_S6", "X4C_S12"), 'regime'] <- 'random'
GE[GE$sample %in% c("X3A_S9", "X3C_S11", "X5A_S2", "X5C_S4", "X6A_S3", "X6C_S10"), 'regime'] <- 'regular'
ggplot(data=GE, mapping=aes(x=regime, y=expression, color=regime)) + geom_boxplot() + ggtitle(Term(GOTERM[goterm]))
```

Among the `r length(genes)` genes in this category, `r sum(topTable(fitmm, coef='regime', number=length(fitmm$F))[genes, 'adj.P.Val'] <= 0.1)`
have an adjusted p-value lower or equal to 0.1. There are too many genes annotated with this function to plot all their stratified expression levels.
I select the most differentially expressed ones.

```{r G2, fig.width=15, fig.height=20}
z <- topTable(fitmm, coef='regime', number=length(fitmm$F))
z <- z[row.names(z) %in% genes, ]   # gene order is preserved:
head(z)
genes <- head(row.names(z), n=50)
GE <- GE[GE$gene %in% genes,]
ggplot(data=GE, mapping=aes(x=regime, y=expression, color=regime)) + geom_boxplot() +
      facet_wrap(~gene) + ggtitle(Term(GOTERM[goterm]))
```

Among the top `r length(genes)` genes analysed from the G protein-coupled receptor signaling pathway,
`r sum(topTable(fitmm, coef='regime', number=length(fitmm$F))[genes, 'logFC'] < 0)`
have a reduced expression level in the unpredictable environment, and
`r sum(topTable(fitmm, coef='regime', number=length(fitmm$F))[genes, 'logFC'] > 0)`
show an increased expression level.

## Cell-matrix adhesion

```{r adhesion}
goterm <- 'GO:0007160'
genes <- names(fullAnnotation[grep(goterm, fullAnnotation, fixed=TRUE)])
genes <- genes[genes %in% row.names(varPart)]
plotPercentBars(varPart[genes,])

topTable(fitmm, coef='regime', number=length(fitmm$F))[genes,]  # not ordered by p-value

z <- as.data.frame(vobjDream$E[genes,])
z$gene <- row.names(z)
GE <- pivot_longer(z, cols=c(1:12), names_to='sample', values_to='expression')
GE$regime <- factor('regular', levels=c('regular', 'random'))
GE[GE$sample %in% c("X1A_S8", "X1C_S1", "X2A_S7", "X2C_S5", "X4A_S6", "X4C_S12"), 'regime'] <- 'random'
ggplot(data=GE, mapping=aes(x=regime, y=expression, color=regime)) + geom_boxplot() + facet_wrap(~gene) + ggtitle(Term(GOTERM[goterm]))
```

Among the top `r length(genes)` genes involved in `r Term(GOTERM[goterm])`, 
`r sum(topTable(fitmm, coef='regime', number=length(fitmm$F))[genes, 'logFC'] < 0)`
have a reduced expression level in the unpredictable environment, and
`r sum(topTable(fitmm, coef='regime', number=length(fitmm$F))[genes, 'logFC'] > 0)`
show an increased expression level.

## Potassium ion transport, carboxylic acid transport, and ion transmembrane transport
Several functions related to ion transport are significant. I wonder to what extent their significance is not
dependent on a common subset of genes.

```{r independence}
goterm_K <- 'GO:0006813'
goterm_COOH <- 'GO:0046942'
goterm_Sulf <- 'GO:0008272'
goterm_Nucl <- 'GO:1901642'
goterm_TM   <- 'GO:0034220'

genes_K    <- names(fullAnnotation[grep(goterm_K,    fullAnnotation, fixed=TRUE)])
genes_COOH <- names(fullAnnotation[grep(goterm_COOH, fullAnnotation, fixed=TRUE)])
genes_Sulf <- names(fullAnnotation[grep(goterm_Sulf, fullAnnotation, fixed=TRUE)])
genes_Nucl <- names(fullAnnotation[grep(goterm_Nucl, fullAnnotation, fixed=TRUE)])
genes_TM   <- names(fullAnnotation[grep(goterm_TM,   fullAnnotation, fixed=TRUE)])

vennDiagram(vennCounts(cbind(Potassium     = row.names(varPart) %in% genes_K,
                             Carboxilic    = row.names(varPart) %in% genes_COOH,
                             Sulfate       = row.names(varPart) %in% genes_Sulf,
                             Nucleoside    = row.names(varPart) %in% genes_Nucl,
                             Transmembrane = row.names(varPart) %in% genes_TM)))
```

There is hardly any overlap among genes involved in the transport of these substances. The only remarkable overlap
happens between `r Term(GOTERM['GO:0006813'])` and `r Term(GOTERM['GO:0034220'])`. But, still the significance of
each term should be quite independent of the significance of any other term. This was expected, because the analysis
of functional enrichment applied algorithms to account for the dependence structure among GO terms. I analyse them
separately, below.

### 

## Cillium movement and cillium assembly

## Proteasome-mediated ubiquitin-dependent protein catabolic process

## Response to oxidative stress

# Molecular functions
## Motor activity

## Peptidase activities: metallocarboxypeptidase activity, metalloendopeptidase activity, serine-type endopeptidase activity and calcium-dependent cysteine-type endopeptidase activity

## oxidoreductase activity, acting on paired donors, with incorporation or reduction of molecular oxygen, reduced ascorbate as one donor, and incorporation of one atom of oxygen

# Session info

# References
