---
title: "Cell functions affected by environmental unpredictability"
author: "J. Ignacio Lucas Lledó"
date: "23/1/2020"
output: html_document
bibliography: interpretation.bib
---

```{r hide, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

On `2020-01-14` I run the analysis of Gene Ontology terms. Here I want to focus on part of those results and make some findings
clearer. I will use results only from gene annotation, and not from annotation of transcripts, which is a somewhat noiser dataset.
I will not look at the functions affected by hatching condition, since it only affects a couple of genes, to be inspected later.
Among all cellular functions significantly enriched with genes differentially expressed among selective regimes, I will focus
on those that are most significant, and detailed enough. For example, I skip terms like *transmembrane transport*, *proteolysis*
or *protein phosphorylation*, because they are too general. Sections below are named for the functional categories that I think
are worth discussing. I will try to contextualize the results. My entry point to the literature is [@GarciaRoger2019].

I need the results of the differential expression analysis (`2020-01-08`) to identify the genes annotated with the significant
GO terms, and to assess the direction of gene expression change between selective regimes.

```{r setup, message=FALSE}
library(variancePartition)
library(GO.db)
library(ggplot2)
library(stringr)
ANNOTATION   <- '../2019-07-26/genes/annotation.txt'
EXPRESSION   <- '../2020-01-08/genes.RData'   # variance partition and the mixed models fitted with dream()
INTERPROSCAN <- '../2019-07-19/functional_annotation.tsv'  # annotation of transcripts
TRANS2GENE   <- 'transcript2gene.txt'   # created in this folder by README.sh from 2019-07-10/transcripts.fa
```

The annotation table includes only the most specific terms that can be assigned to a
gene. The more general, ancestor terms are assumed, of course, but not explicitly
mentioned in the annotation. For the purpose of retrieving all genes annotated to a
term, I need to expand the annotation, to make ancestor terms explicit. It took me a
while to realize that I can do this with the `unlist()` function:

```{r data}
annotation <- read.table(ANNOTATION, col.names = c('gene', 'GOterms'), colClasses = c('character', 'character'))
head(annotation)
annotation.list <- strsplit(annotation$GOterms, split='|', fixed=TRUE)
names(annotation.list) <- annotation$gene
# append can only join two vectors at a time. It works with lists!
allAncestors <- append(as.list(GOBPANCESTOR),
                       append(as.list(GOMFANCESTOR),
                              as.list(GOCCANCESTOR)))

fullAnnotation <- lapply(annotation.list,
                         FUN = function(x){
                           unique(append(x,
                                         unlist(allAncestors[x], use.names=FALSE)))
                         })
load(EXPRESSION)
```

Without any attempt to characterize gene function any further, I will make the assumption
that the more genes in a pathway are expressed, the more active the pathway is. It is,
however, conceivable that some genes are annotated as involved in a function that they
regulate negatively.

Dormant or resting eggs are embryos in a dormant state. RNA was extracted either after
forced diapause or not, but under hatching conditions. That is, a variable portion of
eggs are expected to be leaving the dormancy stage and resuming metabolism, development
and hatching. The portion of eggs getting ready to hatch is known to be larger (actually,
a majority) in populations from regular environments. Thus, some of our results could
resemble those from comparisons between resting and non-resting (amictic) eggs
[@Ziv2017, @Rozema2019].

We know that gene expression levels do not correlate with protein abundances [@Ziv2017].
In dormant embryos, where translation is downregulated [@Ziv2017], this lack of correlation
may be even more important. Many genes found to be expressed in dormant embryos may only
be translated upon exit from dormancy. From this point of view, differences in gene expression
between regimes may not reflect a difference in the active functions at the moment of
RNA extraction, but differences in the original provision.

```{r functions}
# By using 'paste(goterm,collapse='|')', I can retrieve all genes annotated with any of a set of GO terms.
getGenes <- function(goterm, annotation = fullAnnotation, varpart = varPart) {
  genes <- names(annotation[grep(paste(goterm, collapse='|'), annotation)])
  if (! is.null(varpart)) {
    genes <- genes[genes %in% row.names(varpart)]
  }
  return(genes)
}

volcanoPlot <- function(fit=fitmm, Coef='regime', num=length(fitmm$F), term=goterm, genes=getGenes(term)){
  DE <- topTable(fit, coef=Coef, number=num)
  DE$annotation <- factor('other', levels=c('other', Term(GOTERM[term])), ordered=TRUE)
  DE[row.names(DE) %in% genes, 'annotation'] <- Term(GOTERM[term])
  DE <- DE[order(DE$annotation),]
  ggplot(data=DE, mapping=aes(x=logFC, y=-log10(P.Value), color=str_wrap(annotation, width=30))) + geom_point() +
    labs(color='Annotation')
}
```

For adequate interpretation of some categories, I will need to check the original functional annotation
performed with interproscan on `2019-07-19`. From one of the output tables, I select the most informative
columns, potentially including names of functional domains detected in transcripts. I also need the
correspondance between genes and transcripts, which is found in `2019-07-10/transcripts.fa`.

```{r transcripts}
# The file is tab-separated. Some rows miss the last two fields. Some e-values are '-'. And
# functional description strings may include the single quote character.
interpro <- read.table(INTERPROSCAN, sep='\t', fill=TRUE, na.strings=c('-',''), quote='"',
                       colClasses=c('character','NULL','integer','factor','character','character',
                                    'integer','integer','numeric','NULL','NULL','character','character','character'),
                       col.names=c('transcript', NA, 'tsLength','analysis','acc','description', 'start', 'stop',
                                   'evalue', NA, NA,'IPacc','IPdescription','GO'))
interpro$transcript <- substr(interpro$transcript, 1, 14)   # to remove the .p1, .p2... suffix.

ts2gene <- read.table(TRANS2GENE, colClasses=c('character','character'), col.names=c('transcript','gene'))
# Having all the information, I want a function that outputs the interpro annotations of a gene, a
# set of genes, or a GO term.
showFunctions <- function(gene, goterm=NULL, transcriptTable=ts2gene, annotation=interpro){
  if (! is.null(gene)) {
    return(annotation[annotation$transcript %in% transcriptTable[transcriptTable$gene %in% gene, 'transcript'],])
  } else {
    return(showFunctions(getGenes(goterm)))
  }
}
```

# Biological processes
## Nucleotide-excision repair
It is among the most significantly enriched terms, with only 13 genes annotated. In all volcano
plots below, fold changes are computed with the expression level in the unpredictable regime in
the numerator: if the log fold-change is negative, the expression is higher in the regular
environment, and viceversa.

```{r NER}
goterm <- 'GO:0006289'
genes <- getGenes(goterm)
plotPercentBars(varPart[genes[order(-varPart[genes, 'regime'])],
                        c('regime', 'treatment', 'population', 'regime:treatment', 'Residuals')])
volcanoPlot()
```

The expression of NER genes is higher in dormant embryos from populations evolved in
the unpredictable environment. I do not find any similar result in the literature, so
that the significance of this result is open to interpretation. However, it is one of
the most significant and clear results, in addition to potentially being a new
discovery.

## G protein-coupled receptor signaling pathway, and signal transduction

```{r GproteinBars, fig.height=20}
goterm <- 'GO:0007186'
genes <- getGenes(goterm)
plotPercentBars(varPart[genes[order(-varPart[genes, 'regime'])],
                        c('regime', 'treatment', 'population', 'regime:treatment', 'Residuals')])
```

```{r GproteinVolcano}
volcanoPlot()
```

Among the `r length(genes)` genes in this category,
`r sum(topTable(fitmm, coef='regime', number=length(fitmm$F))[genes, 'adj.P.Val'] <= 0.1)`
have an adjusted p-value lower or equal to 0.1. Most of the genes in this pathway significantly
regulated by selective regime, including the most significant ones, are underexpressed in the
unpredictable regime (negative log fold change). Thus, in general we expect dormant eggs from populations
under the unpredictable regime to be less sensitive to this particular pathway of
signal transduction.

```{r signaltransductionBars, fig.width=12}
goterm <- 'GO:0007165'
genes <- getGenes(goterm)
```

```{r signaltransductionVolcano}
volcanoPlot()
```

When we look at all `r length(genes)` genes involved in signal transduction, we still see that
most of them are underexpressed in the unpredictable regime, if differentially expressed at all.
But the pattern is less clear, and some of these genes are also significantly overexpressed in
the unpredictable regime. It is tempting to infer that dormant embryos from unpredictable
environments are less sensitive to the unreliable external estimuli.

## Cell-matrix adhesion

```{r adhesion}
goterm <- 'GO:0007160'
genes <- getGenes(goterm)
plotPercentBars(varPart[genes[order(-varPart[genes, 'regime'])],
                        c('regime', 'treatment', 'population', 'regime:treatment', 'Residuals')])
volcanoPlot()
```

Among the top `r length(genes)` genes involved in `r Term(GOTERM[goterm])`, 
`r sum(topTable(fitmm, coef='regime', number=length(fitmm$F))[genes, 'logFC'] < 0)`
have a reduced expression level in the unpredictable environment, and
`r sum(topTable(fitmm, coef='regime', number=length(fitmm$F))[genes, 'logFC'] > 0)`
show an increased expression level.

Cell-matrix adhesion is likely related to embryonic development. The difference in expression
of genes in this pathway between regimes makes more sense from the perspective of the
expected difference in their current physiological state.

## Potassium ion transport, carboxylic acid transport, and ion transmembrane transport
Several functions related to ion transport are significant. I wonder to what extent their significance is not
dependent on a common subset of genes.

```{r independence}
goterm_K    <- 'GO:0006813'
goterm_COOH <- 'GO:0046942'
goterm_Sulf <- 'GO:0008272'
goterm_Nucl <- 'GO:1901642'
goterm_TM   <- 'GO:0034220'

genes_K    <- getGenes(goterm_K)
genes_COOH <- getGenes(goterm_COOH)
genes_Sulf <- getGenes(goterm_Sulf)
genes_Nucl <- getGenes(goterm_Nucl)
genes_TM   <- getGenes(goterm_TM)

vennDiagram(vennCounts(cbind(Potassium     = row.names(varPart) %in% genes_K,
                             Carboxilic    = row.names(varPart) %in% genes_COOH,
                             Sulfate       = row.names(varPart) %in% genes_Sulf,
                             Nucleoside    = row.names(varPart) %in% genes_Nucl,
                             Transmembrane = row.names(varPart) %in% genes_TM)))
```

There is hardly any overlap among genes involved in the transport of these substances. The only remarkable overlap
happens between `r Term(GOTERM['GO:0006813'])` and `r Term(GOTERM['GO:0034220'])`. But, still the significance of
each term should be quite independent of the significance of any other term. This was expected, because the analysis
of functional enrichment applied algorithms to account for the dependence structure among GO terms. I analyse them
separately, below.

### Potassium ion transport

```{r potassiumBars, fig.height=10}
goterm <- goterm_K
genes <- getGenes(goterm)
plotPercentBars(varPart[genes[order(-varPart[genes, 'regime'])],
                        c('regime', 'treatment', 'population', 'regime:treatment', 'Residuals')])
```

```{r potassiumVolcano}
volcanoPlot()
```

This biological process is defined as "`r Definition(GOTERM[goterm])`". There are `r length(genes)`
in the analysis with this annotation. The pattern is not very consistent.

### Ion transmembrane transport

```{r ionBars, fig.height=10}
goterm <- goterm_TM
genes <- getGenes(goterm)
plotPercentBars(varPart[genes[order(-varPart[genes, 'regime'])],
                        c('regime', 'treatment', 'population', 'regime:treatment', 'Residuals')])
```

```{r ionVolcano}
volcanoPlot()
```

Ion transmembrane transport is an even larger and more general term. One gene sticks out
with a log fold change close to -4. This gene is XLOC_005433. It has homology to the
neurotransmitter-gated ion-channel ligand-binding domain, and to the nicotinic acetylcoline
receptor.

### Nucleoside transmembrane transport

```{r nucleoside}
goterm <- goterm_Nucl
genes  <- getGenes(goterm)
plotPercentBars(varPart[genes[order(-varPart[genes, 'regime'])],
                        c('regime', 'treatment', 'population', 'regime:treatment', 'Residuals')])
volcanoPlot()
```

The most significant gene in this subset is XLOC_035757, an equilibrative nucleoside transporter
slightly more expressed in unpredictable than in predictable environment. Significant only with
FDR 0.091.

### Carboxylic acid transport

```{r COOH}
goterm <- goterm_COOH
genes  <- getGenes(goterm)
plotPercentBars(varPart[genes[order(-varPart[genes, 'regime'])],
                        c('regime', 'treatment', 'population', 'regime:treatment', 'Residuals')])
volcanoPlot()
```

The two most significant genes here, XLOC_017768 and XLOC_017784, are slightly upregulated in the
unpredictable environment. The former seems to be an amino acid permease, and the latter, a phospholipase
A2. Both include a short predicted intrinsically disordered region. 

## Trehalose biosynthetic process

Trehalose is a disaccharide that confers protection against dessication in several organisms [@Clegg1965]. Trehalose
is present in higher concentrations in resting eggs of *B. plicatilis* (which resist dessication) than in asexual eggs
[@Rozema2019]. While trehalose is a very relevant molecule for dormant embryos, it is unclear why genes involved in
its biosynthesis are expressed more in eggs from regular than from unpredictable environments.

```{r trehalose}
goterm <- 'GO:0005992'
genes <- getGenes(goterm)
plotPercentBars(varPart[genes[order(-varPart[genes, 'regime'])],
                        c('regime', 'treatment', 'population', 'regime:treatment', 'Residuals')])
volcanoPlot()
```

## Cell motility, cilium movement and cilium assembly
Adult individuals of *B. plicatilis* have a crown of cilia. Thus, the expression of genes for cilium assembly,
and probably also those for cilium movement and cell motility in the embryo must be related to embryonic
development. They are, thus a sign that the embryos are leaving the dormant state and resuming metabolism.
Under this light, it makes sense that eggs from a regular environment express these genes more than eggs from
the unpredictable environment, because most of the former must be getting ready to hatch. The higher expression
level in predictable environments is a clear pattern in genes involved in all three of these processes.

Dr. Roger has actually observed embryo movement in many of the eggs from
the predictable regime after 30 hours of exposure to hatching conditions (personal
communication).

```{r cillia}
goterm_cellmoti <- 'GO:0048870'   # cell motility
goterm_movement <- 'GO:0003341'   # cilium movement
goterm_assembly <- 'GO:0060271'   # cilium assembly

genes_cellmoti <- getGenes(goterm_cellmoti)
genes_movement <- getGenes(goterm_movement)
genes_assembly <- getGenes(goterm_assembly)

vennDiagram(vennCounts(cbind('Cell motility'    = row.names(varPart) %in% genes_cellmoti,
                             'Cilium movement' = row.names(varPart) %in% genes_movement,
                             'Cilium assembly' = row.names(varPart) %in% genes_assembly)))
```

### Cell motility

```{r cellmoti}
goterm <- goterm_cellmoti
genes  <- getGenes(goterm)
plotPercentBars(varPart[genes[order(-varPart[genes, 'regime'])],
                        c('regime', 'treatment', 'population', 'regime:treatment', 'Residuals')])
volcanoPlot()
```

### Cilium movement

```{r movement}
goterm <- goterm_movement
genes  <- getGenes(goterm)
plotPercentBars(varPart[genes[order(-varPart[genes, 'regime'])],
                        c('regime', 'treatment', 'population', 'regime:treatment', 'Residuals')])
volcanoPlot()
```

### Cilium assembly

```{r assembly, fig.height=10, fig.width=10}
goterm <- goterm_assembly
genes  <- getGenes(goterm)
plotPercentBars(varPart[genes[order(-varPart[genes, 'regime'])],
                        c('regime', 'treatment', 'population', 'regime:treatment', 'Residuals')])
```

```{r assemblyVolcano}
volcanoPlot()
```

## Proteasome-mediated ubiquitin-dependent protein catabolic process

```{r proteasome, fig.height=10, fig.weight=10}
goterm <- 'GO:0043161'
genes  <- getGenes(goterm)
plotPercentBars(varPart[genes[order(-varPart[genes, 'regime'])],
                        c('regime', 'treatment', 'population', 'regime:treatment', 'Residuals')])
```

```{r proteasomeVolcano}
volcanoPlot()
```

It is unclear why preteasome-mediated protein degradation should be more or less active in one or
the other regime. And the pattern is actually not consistent: two genes from this pathway are quite
significantly (but not very much) more expressed in the unpredictable than in the regular environment.
But several other genes in the pathway are less significantly, but to a higher degree underexpressed
in the unpredictable regime. 

## Response to oxidative stress
Oxidative stress is one of the threats faced by dormant embryos, and protection against it is one
of the classic hallmarks of dormancy [@Clark2012, @Ziv2017]. 

```{r oxiestres, fig.height=8}
goterm <- 'GO:0006979'
genes  <- getGenes(goterm)
plotPercentBars(varPart[genes[order(-varPart[genes, 'regime'])],
                        c('regime', 'treatment', 'population', 'regime:treatment', 'Residuals')])
```

```{r oxiestresVolcano}
volcanoPlot()
```

In this case again, the differential expression does not have the same sign in all genes from
this category that are regulated by the selective regime. It looks as if eggs in different
regimes use different sets of genes to protect themselves against oxidative stress.

# Molecular functions
## Motor activity

```{r motorBars, fig.height=15}
goterm <- 'GO:0003774'
genes  <- getGenes(goterm)
plotPercentBars(varPart[genes[order(-varPart[genes, 'regime'])],
                        c('regime', 'treatment', 'population', 'regime:treatment', 'Residuals')])
```

```{r motorVolcano}
volcanoPlot()
```

The three most significant genes here are overexpressed in eggs from the unpredictable
environment, which seems to contrast the results for biological processes of cell motility
and cilium movement. However, `r Term(GOTERM[goterm])` is defined as
`r Definition(GOTERM[goterm])`. Only `r sum(getGenes(goterm) %in% getGenes('GO:0003341'))`
genes involved in `r Term(GOTERM[goterm])` is also involved in `r Term(GOTERM['GO:0003341'])`.

## Peptidase activities: metallocarboxypeptidase activity, metalloendopeptidase activity, serine-type endopeptidase activity and calcium-dependent cysteine-type endopeptidase activity

```{r peptidase}
goterm_metallocarboxy <- 'GO:0004181'
goterm_metalloendopep <- 'GO:0004222'
goterm_serinetypeendo <- 'GO:0004252'
goterm_calciumcystein <- 'GO:0004198'

genes_metallocarboxy <- names(fullAnnotation[grep(goterm_metallocarboxy, fullAnnotation, fixed=TRUE)])
genes_metalloendopep <- names(fullAnnotation[grep(goterm_metalloendopep, fullAnnotation, fixed=TRUE)])
genes_serinetypeendo <- names(fullAnnotation[grep(goterm_serinetypeendo, fullAnnotation, fixed=TRUE)])
genes_calciumcystein <- names(fullAnnotation[grep(goterm_calciumcystein, fullAnnotation, fixed=TRUE)])

vennDiagram(vennCounts(cbind(Metallocarboxy. = row.names(varPart) %in% genes_metallocarboxy,
                             Metalloendo.    = row.names(varPart) %in% genes_metalloendopep,
                             Serine_type = row.names(varPart) %in% genes_serinetypeendo,
                             Cystein_type = row.names(varPart) %in% genes_calciumcystein)))
```

### Metallocarboxypeptidase activity

```{r metallocarboxypeptidase}
goterm <- 'GO:0004181'
genes  <- getGenes(goterm)
plotPercentBars(varPart[genes[order(-varPart[genes, 'regime'])],
                        c('regime', 'treatment', 'population', 'regime:treatment', 'Residuals')])
volcanoPlot()
```

### Serine-type endopeptidase activity

```{r serineBars, fig.height=10}
goterm <- 'GO:0004252'
genes  <- getGenes(goterm)
plotPercentBars(varPart[genes[order(-varPart[genes, 'regime'])],
                        c('regime', 'treatment', 'population', 'regime:treatment', 'Residuals')])
```

```{r serineVolcano}
volcanoPlot()
```

### Calcium-dependent cysteine-type endopeptidase activity

```{r calciumBars, fig.height=8}
goterm <- 'GO:0004198'
genes  <- getGenes(goterm)
plotPercentBars(varPart[genes[order(-varPart[genes, 'regime'])],
                        c('regime', 'treatment', 'population', 'regime:treatment', 'Residuals')])
```

```{r calciumVolcano}
volcanoPlot()
```

### Metalloendopeptidase activity

```{r metalloendoBars, fig.height=10}
goterm <- 'GO:0004222'
genes  <- getGenes(goterm)
plotPercentBars(varPart[genes[order(-varPart[genes, 'regime'])],
                        c('regime', 'treatment', 'population', 'regime:treatment', 'Residuals')])
```

```{r metalloendoVolcano}
volcanoPlot()
```

## oxidoreductase activity, acting on paired donors, with incorporation or reduction of molecular oxygen, reduced ascorbate as one donor, and incorporation of one atom of oxygen

```{r oxidoreductaseBars}
goterm <- 'GO:0016715'
genes  <- getGenes(goterm)
plotPercentBars(varPart[genes[order(-varPart[genes, 'regime'])],
                        c('regime', 'treatment', 'population', 'regime:treatment', 'Residuals')])
volcanoPlot()
```


# Session info
```{r sessioninfo}
sessionInfo()
```

# References
