---
title: "Detailed analysis of some candidate genes"
author: "J. Ignacio Lucas Lledó"
date: "27/10/2020"
output: html_document
---

```{r setup, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(edgeR)
library(variancePartition)
library(topGO)
library(ggplot2)
load('../2020-01-08/genes.RData')
load('../2020-01-14/Enrichment.genes.regime.RData')
```

```{r message=FALSE, warning=FALSE}
GO.list <- as.list(GOTERM)
vitelline.GO <- grep("vitelline", sapply(GO.list, function(x) Term(x)), value=TRUE)
fattyacid.GO <- grep("fatty acid synthase", sapply(GO.list, function(x) Term(x)), value=TRUE)

vitelline.genes <- unique(unlist(
  sapply(list(GOdata.BP, GOdata.CC, GOdata.MF), function(x) {
    genesInTerm(x, names(vitelline.GO))
  })
))

fattyacid.genes <- unique(unlist(
  sapply(list(GOdata.BP, GOdata.CC, GOdata.MF), function(x) {
    genesInTerm(x, names(fattyacid.GO))
  })
))
```

There are `r length(fattyacid.genes)` genes in the data set with annotations related to the
fatty acid synthase, and `r length(vitelline.genes)` genes associated with vitelline.

```{r }
NumOfGenes <- dim(vobjDream)[1]
topTable(fitmm, coef='regime', number=NumOfGenes, sort.by='p')[fattyacid.genes,]
MF.all <- GenTable(GOdata.MF, elim=MF.elim, weight01=MF.weight01, lea=MF.lea, orderBy='weight01', ranksOf='elim', topNodes=length(usedGO(GOdata.MF)))
MF.all[MF.all$GO.ID %in% names(fattyacid.GO),]

```

# Conclusions
In our data set there is no gene with any annotation related to the vitelline outer membrane.
It is possible that such genes did not pass the abundance threshold applied before differential
expression.

There are 5 genes associated with the "fatty acid synthase activity" (molcular function), and
one of them is significantly differentially expressed (raw $p$ value 0.00513479). It's fold change
is positive (0.36), indicating higher expression levels in the random environment. 


# Session Information
```{r}
sessionInfo()
```
